---
title: "R Notebook"
output: html_notebook
---

```{r, import libraries}
library(igraph)
library(SingleCellExperiment)
library(S4Vectors)
library(stringr)
library(DT)

library(dplyr)
library(tidyr)
library(mclust)
library(ggplot2)
library(RColorBrewer)
library(scater)
library(Rphenoannoy)
set.seed(101100)
```

```{r, Set wd and load data}
#set working directory 
wd <-dirname(getwd())

data_folder <-(file.path(wd,"sce_objects","tumour_non-tumour_from_TMA"))
all.cells <- readRDS(file=file.path(data_folder, "all-cells_merged.rds"))

all.tumour <- readRDS(file=file.path(data_folder, "merge_all-TUMOUR.rds"))
```


```{r, Define good markers, echo=F, warning=F, message=FALSE}
all.marker <-rownames(all.cells)

bad.marker <- c("Iridium_191","Iridium_193","Cadherin-6","Histone H3") 
good.marker <- all.marker[!all.marker %in% bad.marker]
print(good.marker)
```

```{r}
all.cells$Tma_ac <- paste(all.cells$TMA, all.cells$acID, sep="_")
length(unique(all.cells$Tma_ac))
```


```{r, sub tumour final, warning=F, message=F, echo=F, eval=F}
#split cells by ImageNumber
n_cells_dr <- 100
cs <- split(seq_len(ncol(all.cells)), all.cells$Tma_ac)
length(unique(all.cells$Tma_ac))
#sample 'n_cells_dr' per ImageNumber
cs <- unlist(lapply(cs, function(u) 
  sample(u, min(n_cells_dr, length(u)))))
#sub.test <- sce[,cs]
all.cells.sub <- all.cells[,cs]

#calculate percentage of cells subsetting results in
p<-dim(assay(all.cells.sub))[2]/dim(assay(all.cells))[2]*100

#results in % of all cells
print(paste("Subsetting results in a total number of",dim(assay(
  all.cells.sub))[2],"cells, which corresponds to",round(p,digits=2), "% of all cells of the original sce." ))

saveRDS(all.cells.sub, file=file.path(data_folder, paste("all_cells_sub.rds")))
```

```{r, load subset, message=FALSE, warning=FALSE, echo=FALSE}
all.cells.sub <- readRDS(file=file.path(data_folder, paste("all_cells_sub.rds")))
```

```{r, calculate umap, warning=F, message=F, echo=F, eval=FALSE}
p <-c(10,50,100)

for(i in p){
all.cells.sub <- runUMAP(all.cells.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("UMAP_p", i),
                 #use_dimred="PCA_20",
                 n_neighbors = i)
saveRDS(all.cells.sub, file=file.path(data_folder, paste("all_cells_sub.rds")))

}
saveRDS(all.cells.sub, file=file.path(data_folder, paste("all_cells_sub.rds")))


for(i in p){
all.cells.sub <- runTSNE(all.cells.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("tSNE_p", i),
                 #use_dimred="PCA_20",
                 perplexity = i)
saveRDS(all.cells.sub, file=file.path(data_folder, paste("all_cells_sub.rds")))
}
```


**UMAP with Tumour markers**
```{r,plot umap tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.cells.sub)$`UMAP_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.cells.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```

**tsne with Tumour markers**
```{r,plot tsne tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.cells.sub)$`tSNE_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.cells.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```


```{r, plot umap tumour cluster, fig.width=6, fig.height=4, warning=F, message=F, echo=F}
cluster <- "mclust"
p <-plotReducedDim(all.cells.sub, "UMAP_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="UMAP-1", y="UMAP-2", title="UMAP") 
plot(p)
##ggsave(filename=file.path(plot_folder, paste("sub_Fibro_CAF-Clusters_UMAP_p50.png",sep="")), plot=p)

cluster <- "mclust"
p <-plotReducedDim(all.cells.sub, "tSNE_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="tSNE-1", y="tSNE-2", title="tSNE") 
plot(p)
```
##TUMOUR
```{r}
all.tumour <- readRDS(file=file.path(data_folder, "merge_all-TUMOUR.rds"))

all.tumour$TMA <- paste(all.tumour$TmaID, all.tumour$TmaBlock, sep="")

all.tumour$Tma_ac <- paste(all.tumour$TMA, all.tumour$acID, sep="_")
length(unique(all.tumour$Tma_ac))
head(colData(all.tumour))
```


```{r, sub tumour final, warning=F, message=F, echo=F, eval=F}
#split cells by ImageNumber
n_cells_dr <- 100
cs <- split(seq_len(ncol(all.tumour)), all.tumour$Tma_ac)
length(unique(all.tumour$Tma_ac))
#sample 'n_cells_dr' per ImageNumber
cs <- unlist(lapply(cs, function(u) 
  sample(u, min(n_cells_dr, length(u)))))
#sub.test <- sce[,cs]
all.tumour.sub <- all.tumour[,cs]

#calculate percentage of cells subsetting results in
p<-dim(assay(all.tumour.sub))[2]/dim(assay(all.tumour))[2]*100

#results in % of all cells
print(paste("Subsetting results in a total number of",dim(assay(
  all.tumour.sub))[2],"cells, which corresponds to",round(p,digits=2), "% of all cells of the original sce." ))

saveRDS(all.tumour.sub, file=file.path(data_folder, paste("all_tumour_sub.rds")))
```

```{r, load subset, message=FALSE, warning=FALSE, echo=FALSE}
all.tumour.sub <- readRDS(file=file.path(data_folder, paste("all_tumour_sub.rds")))
```

```{r, calculate umap, warning=F, message=F, echo=F, eval=FALSE}
p <-c(10,50,100)

for(i in p){
all.tumour.sub <- runUMAP(all.tumour.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("UMAP_p", i),
                 #use_dimred="PCA_20",
                 n_neighbors = i)
saveRDS(all.tumour.sub, file=file.path(data_folder, paste("all_tumour_sub.rds")))

}
saveRDS(all.tumour.sub, file=file.path(data_folder, paste("all_tumour_sub.rds")))

p=50
for(i in p){
all.tumour.sub <- runTSNE(all.tumour.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("tSNE_p", i),
                 #use_dimred="PCA_20",
                 perplexity = i)
saveRDS(all.tumour.sub, file=file.path(data_folder, paste("all_tumour_sub.rds")))
}
```


**UMAP with Tumour markers**
```{r,plot umap tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.tumour.sub)$`UMAP_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.tumour.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```

**tsne with Tumour markers**
```{r,plot tsne tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.tumour.sub)$`tSNE_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.tumour.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```




```{r Clustering using Rphenoannoy, fig.width=25, fig.height=12}
hmcol<-rev(brewer.pal(11,"RdBu"))
rdylbu <-rev(brewer.pal(11,"RdYlBu"))

#marker.clustering <- good.marker[!good.marker %in% "Pan Cytokeratin + Keratin Epithelial"]
#cellnb <- 250000
#subsample <-  colnames(sce)[sample(length(colnames(sce)), cellnb)]
#sce_sub <- sce[, colnames(sce) %in% subsample]

#set ks for clustering
#i <- 100
k <- c(30,35,40)
#i<-10
for (i in k) {
    all.tumour$RPmembership <- factor(Rphenoannoy(data = t(assay(all.tumour[rownames(all.tumour) %in% good.marker,],"c_counts_asinh")), k = i)[[2]]$membership)
    cluster <- paste0("rp_tumour_k",i)
  colnames(colData(all.tumour))[which(names(colData(all.tumour)) == "RPmembership")] <- paste0(cluster)
  agg_sce <-aggregateAcrossCells(all.tumour, ids=all.tumour[[cluster]], average=TRUE, use_exprs_values="c_counts_asinh_scaled" )

 
    scater::plotHeatmap(agg_sce,
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              colour_columns_by = c(cluster),
              main=paste0("Heatmap tumour cells, ",cluster))
    
  #save sce clustering
  saveRDS(all.tumour, file=file.path(data_folder, "merge_all-TUMOUR_workingfile.rds"))
}
all.tumour <-readRDS( file=file.path(data_folder, "merge_all-TUMOUR_workingfile.rds"))

```

```{r, add clustering results to subset tumour}
rp_df <- data.frame("CellID"=all.tumour$CellID, "rp_tumour_k30"=all.tumour$rp_tumour_k30,
                    "rp_tumour_k35"=all.tumour$rp_tumour_k35,
                    "rp_tumour_k40"=all.tumour$rp_tumour_k40 )

cur_DF <- as_tibble(colData(all.tumour.sub)) %>% left_join(rp_df, by = "CellID") %>% DataFrame()
all.equal(paste(cur_DF$ImageNumber, cur_DF$CellNumber), paste(all.tumour.sub$ImageNumber, all.tumour.sub$CellNumber))

colData(all.tumour.sub) <- cur_DF
rownames(colData(all.tumour.sub)) <-all.tumour.sub$CellID
```
```{r,define tumour non tumour from clustering}
is.nt <-c(39,12,53,104,47,20,1,97,52,110,150,128,71,32,147,34,138,133,101,8,115)#145
all.tumour$tumour_nontumour <-ifelse(all.tumour$rp_tumour_k40 %in% is.nt, "non_tumour","tumour")
table(all.tumour$tumour_nontumour)

saveRDS(all.tumour, file=file.path(data_folder, "merge_all-TUMOUR_workingfile.rds"))

all.tumour.sub$tumour_nontumour <-ifelse(all.tumour.sub$rp_tumour_k40 %in% is.nt, "non_tumour","tumour")
saveRDS(all.tumour.sub, file=file.path(data_folder, paste("all_tumour_sub_workingfile.rds")))
table(all.tumour$tumour_nontumour)
```


```{r, plot umap all.tumour cluster, fig.width=6, fig.height=4, warning=F, message=F, echo=F}
cluster <- "tumour_nontumour"
p <-plotReducedDim(all.tumour.sub, "UMAP_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="UMAP-1", y="UMAP-2", title="UMAP") 
plot(p)
##ggsave(filename=file.path(plot_folder, paste("sub_Fibro_CAF-Clusters_UMAP_p50.png",sep="")), plot=p)

cluster <- "rp_tumour_k40"
p <-plotReducedDim(all.tumour.sub, "UMAP_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="UMAP-1", y="UMAP-2", title="UMAP") 
plot(p)
#cluster <- "mclust"
p <-plotReducedDim(all.tumour.sub, "tSNE_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="tSNE-1", y="tSNE-2", title="tSNE") 
plot(p)
```

```{r, fig.width=25, fig.height=20}
agg_sce <-aggregateAcrossCells(all.tumour[, all.tumour$tumour_nontumour=="non_tumour"], ids=all.tumour[, all.tumour$tumour_nontumour=="non_tumour"][[cluster]], average=TRUE, use_exprs_values="c_counts_asinh_scaled" )

scater::plotHeatmap(agg_sce,
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              colour_columns_by = c(cluster),
              main=paste0("Heatmap tumour cells, ",cluster))

scater::plotHeatmap(all.tumour[, all.tumour$rp_tumour_k40==47],
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              colour_columns_by = c(cluster),
              main=paste0("Heatmap tumour cells, ",cluster))
```

```{r, subset tumour into tumour_tumour and tumour_non tumour}
tumour_tumour <- all.tumour[, all.tumour$tumour_nontumour=="tumour"]
tumour_NONtumour <- all.tumour[, all.tumour$tumour_nontumour=="non_tumour"]

saveRDS(tumour_tumour, file=file.path(data_folder, "tumour", paste("merge_all-tumour_TUMOUR.rds")))
saveRDS(tumour_NONtumour, file=file.path(data_folder, "tumour", paste("merge_all-tumour_NONtumour.rds")))

tumour_tumour <-readRDS( file=file.path(data_folder, "tumour", paste("merge_all-tumour_TUMOUR.rds")))

```

##NON TUMOUR
```{r}
all.NONtumour <- readRDS(file=file.path(data_folder, "merge_all-NONtumour.rds"))

all.NONtumour$TMA <- paste(all.NONtumour$TmaID, all.NONtumour$TmaBlock, sep="")

all.NONtumour$Tma_ac <- paste(all.NONtumour$TMA, all.NONtumour$acID, sep="_")
length(unique(all.NONtumour$Tma_ac))
head(colData(all.NONtumour))
```


```{r, sub non tumour, warning=F, message=F, echo=F, eval=F}
#split cells by ImageNumber
n_cells_dr <- 100
cs <- split(seq_len(ncol(all.NONtumour)), all.NONtumour$Tma_ac)
length(unique(all.NONtumour$Tma_ac))
#sample 'n_cells_dr' per ImageNumber
cs <- unlist(lapply(cs, function(u) 
  sample(u, min(n_cells_dr, length(u)))))
#sub.test <- sce[,cs]
all.NONtumour.sub <- all.NONtumour[,cs]

#calculate percentage of cells subsetting results in
p<-dim(assay(all.NONtumour.sub))[2]/dim(assay(all.NONtumour))[2]*100

#results in % of all cells
print(paste("Subsetting results in a total number of",dim(assay(
  all.NONtumour.sub))[2],"cells, which corresponds to",round(p,digits=2), "% of all cells of the original sce." ))

saveRDS(all.NONtumour.sub, file=file.path(data_folder, paste("all_NONtumour_sub.rds")))
```

```{r, load subset, message=FALSE, warning=FALSE, echo=FALSE}
all.NONtumour.sub <- readRDS(file=file.path(data_folder, paste("all_NONtumour_sub.rds")))
```

```{r, calculate umap, warning=F, message=F, echo=F, eval=FALSE}
p <-c(10,50,100)

for(i in p){
all.NONtumour.sub <- runUMAP(all.NONtumour.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("UMAP_p", i),
                 #use_dimred="PCA_20",
                 n_neighbors = i)
saveRDS(all.NONtumour.sub, file=file.path(data_folder, paste("all_NONtumour_sub.rds")))

}
saveRDS(all.NONtumour.sub, file=file.path(data_folder, paste("all_NONtumour_sub.rds")))

p=50
for(i in p){
all.NONtumour.sub <- runTSNE(all.NONtumour.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("tSNE_p", i),
                 #use_dimred="PCA_20",
                 perplexity = i)
saveRDS(all.NONtumour.sub, file=file.path(data_folder, paste("all_NONtumour_sub.rds")))
}
```


**UMAP with Tumour markers**
```{r,plot umap tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.NONtumour.sub)$`UMAP_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.NONtumour.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```

**tsne with Tumour markers**
```{r,plot tsne tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.NONtumour.sub)$`tSNE_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.NONtumour.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```




```{r Clustering using Rphenoannoy, fig.width=15, fig.height=15}
hmcol<-rev(brewer.pal(11,"RdBu"))

#marker.clustering <- good.marker[!good.marker %in% "Pan Cytokeratin + Keratin Epithelial"]
#cellnb <- 250000
#subsample <-  colnames(sce)[sample(length(colnames(sce)), cellnb)]
#sce_sub <- sce[, colnames(sce) %in% subsample]

#set ks for clustering
#i <- 100
k <- c(30)
#i<-30
for (i in k) {
    all.NONtumour$RPmembership <- factor(Rphenoannoy(data = t(assay(all.NONtumour[rownames(all.NONtumour) %in% good.marker,],"c_counts_asinh")), k = i)[[2]]$membership)
    cluster <- paste0("rp_NONtumour_k",i)
  colnames(colData(all.NONtumour))[which(names(colData(all.NONtumour)) == "RPmembership")] <- paste0(cluster)
  agg_sce <-aggregateAcrossCells(all.NONtumour, ids=all.NONtumour[[cluster]], average=TRUE, use_exprs_values="c_counts_asinh_scaled" )

 
    scater::plotHeatmap(agg_sce,
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              colour_columns_by = c(cluster),
              main=paste0("Heatmap NON Tumour cells, ",cluster))
    
  #save sce clustering
  saveRDS(all.NONtumour, file=file.path(data_folder, paste("merge_all-NONtumour_workingfile.rds",sep="")))
}
```

```{r}

colnames(colData(all.NONtumour))[!colnames(colData(all.NONtumour)) %in%colnames(colData(tumour_tumour))]
colnames(colData(all.NONtumour))

tumour_NONtumour$rp_tumour_k30 <-NULL
tumour_NONtumour$rp_tumour_k35 <-NULL
tumour_NONtumour$rp_tumour_k40 <-NULL

tumour_NONtumour$tumour_nontumour <-NULL
all.NONtumour$rp_NONtumour_k30 <-NULL
all.NONtumour.final <- cbind(all.NONtumour, tumour_NONtumour)

all.NONtumour.final$tumour_nontumour <-"non tumour"
saveRDS(all.NONtumour.final, file=file.path(data_folder,"non_tumour", paste("merge_all-NONtumour_final.rds",sep="")))


tumour_tumour$rp_tumour_k30 <-NULL
tumour_tumour$rp_tumour_k35 <-NULL
tumour_tumour$rp_tumour_k40 <-NULL

tumour_tumour$tumour_nontumour <-NULL
tumour_tumour$tumour_nontumour <-"tumour"

colnames(colData(all.NONtumour.final))[!colnames(colData(all.NONtumour.final)) %in%colnames(colData(tumour_tumour))]

all.final <- cbind(tumour_tumour,all.NONtumour.final)
saveRDS(all.final, file=file.path(data_folder, paste("merge_all-cells-tumourVSnontumour_final.rds",sep="")))
```
#FINAL NON TUMOUR CLUSTERING to distinguish between stroma / immune (and eventual left over tumour cells (which would need to be remerged to the tumour_tumour dataset))
```{r}
wd <-dirname(getwd())

data_folder <-(file.path(wd,"sce_objects","tumour_non-tumour_from_TMA"))
all.NONtumour.final <- readRDS(file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final.rds"))

all.NONtumour.final$Tma_ac %>% unique() %>% length()
ac_clinical$Tma_ac %>% unique() %>% length()

unique(all.NONtumour.final$Tma_ac)[!unique(all.NONtumour.final$Tma_ac) %in%ac_clinical$Tma_ac]
unique(ac_clinical$Tma_ac)[!unique(ac_clinical$Tma_ac) %in%all.NONtumour.final$Tma_ac]
```


```{r, Define good markers, echo=F, warning=F, message=FALSE}
all.marker <-rownames(all.NONtumour.final)

bad.marker <- c("Iridium_191","Iridium_193","Cadherin-6","Histone H3") 
good.marker <- all.marker[!all.marker %in% bad.marker]
print(good.marker)
```

```{r, sub non tumour, warning=F, message=F, echo=F, eval=F}
#split cells by ImageNumber
n_cells_dr <- 100
cs <- split(seq_len(ncol(all.NONtumour.final)), all.NONtumour.final$Tma_ac)
length(unique(all.NONtumour.final$Tma_ac))
#sample 'n_cells_dr' per ImageNumber
cs <- unlist(lapply(cs, function(u) 
  sample(u, min(n_cells_dr, length(u)))))
#sub.test <- sce[,cs]
all.NONtumour.final.sub <- all.NONtumour.final[,cs]

#calculate percentage of cells subsetting results in
p<-dim(assay(all.NONtumour.final.sub))[2]/dim(assay(all.NONtumour.final))[2]*100

#results in % of all cells
print(paste("Subsetting results in a total number of",dim(assay(
  all.NONtumour.final.sub))[2],"cells, which corresponds to",round(p,digits=2), "% of all cells of the original sce." ))

saveRDS(all.NONtumour.final.sub, file=file.path(data_folder,"non_tumour", paste("all_NONtumour-final_sub.rds")))
```

```{r, load subset, message=FALSE, warning=FALSE, echo=FALSE}
all.NONtumour.final.sub <- readRDS(file=file.path(data_folder, paste("all_NONtumour-final_sub.rds")))
```

```{r, calculate umap, warning=F, message=F, echo=F, eval=FALSE}
p <-c(50)

for(i in p){
all.NONtumour.final.sub <- runUMAP(all.NONtumour.final.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("UMAP_p", i),
                 #use_dimred="PCA_20",
                 n_neighbors = i)
saveRDS(all.NONtumour.final.sub, file=file.path(data_folder,"non_tumour", paste("all_NONtumour-final_sub.rds")))

}
saveRDS(all.NONtumour.final.sub, file=file.path(data_folder,"non_tumour", paste("all_NONtumour-final_sub.rds")))

p=50
for(i in p){
all.NONtumour.final.sub <- runTSNE(all.NONtumour.final.sub,
                 exprs_values = "c_counts_asinh",
                 name = paste0("tSNE_p", i),
                 #use_dimred="PCA_20",
                 perplexity = i)
saveRDS(all.NONtumour.final.sub, file=file.path(data_folder,"non_tumour", paste("all_NONtumour-final_sub.rds")))
}
```


**UMAP with Tumour markers**
```{r,plot umap tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.NONtumour.final.sub)$`UMAP_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.NONtumour.final.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```

**tsne with Tumour markers**
```{r,plot tsne tumour marker, fig.width=12, fig.height=8, echo=F, message=FALSE,warning=FALSE}
dat <-as.data.frame(reducedDims(all.NONtumour.final.sub)$`tSNE_p50`)
dat$cell <- rownames(dat)
dat.counts <-as.data.frame(t((assay(all.NONtumour.final.sub,"c_counts_asinh_scaled"))))
dat.counts$cell <- rownames(dat.counts)
dat.all <-merge(dat, dat.counts, by.x="cell", by.y="cell")

dat.all.long <-pivot_longer(dat.all, cols = colnames(dat.all)[colnames(dat.all) %in% good.marker], names_to = "target", values_to = "counts")


p <-dat.all.long %>%
  ggplot(aes(x=V1, y=V2, color=counts))+
  facet_wrap(~target, scales = "free", ncol = 6)+
  geom_point(alpha=0.5, size=0.2)+
  scale_color_gradientn(colours=rev(brewer.pal(11, 'Spectral')), name='Counts')+
  ggtitle('')+
  theme(strip.background = element_blank(),
        axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        plot.background=element_blank(),
         strip.text = element_text(size=8)) 

##ggsave(filename=file.path(plot_folder, paste("sub_fibro_fibro-Marker_UMAP_p50_NEW.png",sep="")), plot=p, width=16, height=10)
plot(p)
```




```{r Clustering using Rphenoannoy, fig.width=20, fig.height=15}
hmcol<-rev(brewer.pal(11,"RdBu"))
rdylbu <-rev(brewer.pal(11,"RdYlBu"))

#marker.clustering <- good.marker[!good.marker %in% "Pan Cytokeratin + Keratin Epithelial"]
#cellnb <- 250000
#subsample <-  colnames(sce)[sample(length(colnames(sce)), cellnb)]
#sce_sub <- sce[, colnames(sce) %in% subsample]

#set ks for clustering
#i <- 100
k <- c(30,35,40)
#i<-30
for (i in k) {
  #  all.NONtumour.final$RPmembership <- factor(Rphenoannoy(data = t(assay(all.NONtumour.final[rownames(all.NONtumour.final) %in% good.marker,],"c_counts_asinh")), k = i)[[2]]$membership)
   cluster <- paste0("rp_NONtumour-final_k",i)
#  colnames(colData(all.NONtumour.final))[which(names(colData(all.NONtumour.final)) == "RPmembership")] <- paste0(cluster)
  agg_sce <-aggregateAcrossCells(all.NONtumour.final, ids=all.NONtumour.final[[cluster]], average=TRUE, use_exprs_values="c_counts_asinh_scaled" )

 
    scater::plotHeatmap(agg_sce,
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              colour_columns_by = c(cluster),
              main=paste0("Heatmap final NON Tumour cells, ",cluster))
    
  #save sce clustering
 # saveRDS(all.NONtumour.final, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile.rds"))
}
```


```{r}
is.tumour <- c(35, 73,66,75,51,53)
is.immune <- c(59,48,70,58,69,23,50,44,64,30,6,60,12,56,27,47,31,4,29,62,65,26,52,74,39,15,17,22,38,39,21)
is.stroma <- c(53,16,45,46,35,3,7,28,8,18,24,11,36,20,13,10,41,25,55,54,68,71,42,40,61,57,19,51,14,34,63,43,67,37,33,32,1,2,72, 49)
is.undef <- c(9,5)

all.NONtumour.final$cell_category[all.NONtumour.final$`rp_NONtumour-final_k30` %in% is.tumour] <- "tumour"
all.NONtumour.final$cell_category[all.NONtumour.final$`rp_NONtumour-final_k30` %in% is.immune] <- "immune"
all.NONtumour.final$cell_category[all.NONtumour.final$`rp_NONtumour-final_k30` %in% is.stroma] <- "stroma"
all.NONtumour.final$cell_category[all.NONtumour.final$`rp_NONtumour-final_k30` %in% is.undef] <- "undefined"

table(all.NONtumour.final$cell_category)

```

```{r, add clustering results to subset tumour}
rp_df <- data.frame("CellID"=all.NONtumour.final$CellID, "rp_NONtumour-final_k30"=all.NONtumour.final$`rp_NONtumour-final_k30`,
                    "cell_category"=all.NONtumour.final$cell_category)

cur_DF <- as_tibble(colData(all.NONtumour.final.sub)) %>% left_join(rp_df, by = "CellID") %>% DataFrame()
all.equal(paste(cur_DF$ImageNumber, cur_DF$CellNumber), paste(all.NONtumour.final.sub$ImageNumber, all.NONtumour.final.sub$CellNumber))

colData(all.NONtumour.final.sub) <- cur_DF
rownames(colData(all.NONtumour.final.sub)) <-all.NONtumour.final.sub$CellID

cluster <- "cell_category"
p <-plotReducedDim(all.NONtumour.final.sub, "UMAP_p50", colour_by=paste(cluster), text_by = paste(cluster), point_size=0.5,)+labs(x="UMAP-1", y="UMAP-2", title="UMAP") 
plot(p)
```


```{r, fig.width=20, fig.height=10}
cluster <- paste0("rp_NONtumour-final_k",30)
agg_sce <-aggregateAcrossCells(all.NONtumour.final, ids=all.NONtumour.final[[cluster]], average=TRUE, use_exprs_values="c_counts_asinh_scaled" )

 
    scater::plotHeatmap(agg_sce,
              features = good.marker,
              #features = all.marker,
              #features = marker.clustering,
              exprs_values = "c_counts_asinh_scaled",
              symmetric = FALSE,
              zlim=c(0,1),
              color=rdylbu,
              sortrowss=FALSE,
              show_colnames=TRUE,
              color_columns_by = c("cell_category"),
              main=paste0("Heatmap final NON Tumour cells, ",cluster))
```
```{r}
saveRDS(all.NONtumour.final, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile.rds"))

all.NONtumour.final_TUMOUR <- all.NONtumour.final[,all.NONtumour.final$cell_category=="tumour"]
all.NONtumour.final_IMMUNE <- all.NONtumour.final[,all.NONtumour.final$cell_category=="immune"]
all.NONtumour.final_STROMA <- all.NONtumour.final[,all.NONtumour.final$cell_category=="stroma"]
all.NONtumour.final_UNDEF <- all.NONtumour.final[,all.NONtumour.final$cell_category=="undefined"]

saveRDS(all.NONtumour.final_TUMOUR, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile_TUMOUR.rds"))
saveRDS(all.NONtumour.final_IMMUNE, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile_IMMUNE.rds"))
saveRDS(all.NONtumour.final_STROMA, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile_STROMA.rds"))
saveRDS(all.NONtumour.final_UNDEF, file=file.path(data_folder,"non_tumour", "merge_all-NONtumour_final_workingfile_UNDEFINED.rds"))

all.NONtumour.final_minusT <- all.NONtumour.final[,all.NONtumour.final$cell_category!="tumour"]

```

```{r, merge non.tumour_TUMOUR back to tumour_tumour}
colnames(colData(all.NONtumour.final_TUMOUR))[!colnames(colData(all.NONtumour.final_TUMOUR)) %in%colnames(colData(tumour_tumour))]
colnames(colData(tumour_tumour))[!colnames(colData(tumour_tumour)) %in%colnames(colData(all.NONtumour.final_TUMOUR))]

all.NONtumour.final_TUMOUR$cell_category <-NULL
all.NONtumour.final_TUMOUR$`rp_NONtumour-final_k30` <-NULL
all.NONtumour.final_TUMOUR$`rp_NONtumour-final_k35` <-NULL
all.NONtumour.final_TUMOUR$`rp_NONtumour-final_k40` <-NULL

tumour_tumour$rp_tumour_k30 <-NULL
tumour_tumour$rp_tumour_k35 <-NULL
tumour_tumour$rp_tumour_k40 <-NULL

tumour.final <-cbind(tumour_tumour, all.NONtumour.final_TUMOUR)

saveRDS(tumour.final, file=file.path(data_folder,"tumour", "FINAL_merge_all_tumour-final_after-nontumourclustering.rds"))



colnames(colData(all.NONtumour.final_minusT))[!colnames(colData(all.NONtumour.final_minusT)) %in%colnames(colData(tumour.final))]
colnames(colData(tumour.final))[!colnames(colData(tumour.final)) %in%colnames(colData(all.NONtumour.final_minusT))]

tumour.final$cell_category <-"tumour"
all.NONtumour.final_minusT$`rp_NONtumour-final_k30` <-NULL
all.NONtumour.final_minusT$`rp_NONtumour-final_k35` <-NULL
all.NONtumour.final_minusT$`rp_NONtumour-final_k40` <-NULL

all.category <- cbind(tumour.final,all.NONtumour.final_minusT)
all.category$cat <- ifelse(all.category$cell_category=="tumour","tumour","non_tumour")

all.category
```

