---
title: "R Notebook - Analysis T cells"
output:
  html_document:
    df_print: paged
---

```{r, import libraries, echo=F, warnings=F, message=F}
library(igraph)
library(SingleCellExperiment)
library(S4Vectors)
library(stringr)
library(DT)
library(dplyr)
library(tidyr)
library(mclust)
library(ggplot2)
library(RColorBrewer)
library(scater)
library(Rphenoannoy)
library(factoextra)
library(cluster)
library(dendextend)
library(ggthemes)
library(ggpubr)
library(dplyr)
#library(tidyverse)
library(RColorBrewer)
library(pals)
library(qwraps2)
library(table1)
library(SingleCellExperiment)
library(tidyr)
library(scater)
library(ggridges)
library(viridis)
library(viridisLite)
library(ggplot2)
library(data.table)
library(CATALYST)
library(gridExtra)
library(Rphenograph)
library(ComplexHeatmap)
library(CATALYST)
library(scales)
library(survival)
library(broom)
library(pheatmap)
library(FlowSOM)
library(Seurat)
library(Rphenoannoy)
library(dplyr)
library(data.table)
library(ggthemes)
library(diffcyt)
library(edgeR)
library(rstatix)
library(dendextend)
library(ggdendro)
library(dendextend)
library(FactoMineR)
library(factoextra)
library(survminer)
library(corrplot)
library(rstatix)
library(graphics)
library(cowplot)
library(cluster)
library(glmnet)
library(fastDummies)
library(ggsci)
library(FactoMineR)
library(factoextra)
set.seed(101100)
```



```{r, Set wd and load data, echo=F, warnings=F, message=F}
#set working directory 
wd <-dirname(getwd())

data_folder <-(file.path(wd,"sce_objects","immune","Tcell"))

#RAW
#tcell.sce <- readRDS(file=file.path(data_folder, "FINAL_Tcells-minusImmune_workingfile.rds"))
#tcell.sce <- readRDS(file=file.path(data_folder, "FINAL_Analysis_Tcell-clinicaldata_area_RAW.rds"))

#workingfile
#saveRDS(tcell.sce, file=file.path(data_folder, "FINAL_Analysis_Tcell-clinicaldata_area_RAW.rds"))
#saveRDS(tcell.sce, file=file.path(data_folder, "FINAL_Analysis_Tcell-clinicaldata_workingfile.rds"))

#clinical.data
data_folder <-(file.path(wd,"sce_objects","final objects with categories","FINAL"))
tcell.sce <- readRDS(file=file.path(data_folder, "TCELL-only_CLINICAL-DATA_FILTERED.rds"))
tcell.sce$DX.name[is.na(tcell.sce$DX.name)]<-"NA"

```

```{r, load clinical data, echo=F, warnings=F, message=F}
clinical.data <- read.csv(file=file.path(wd,"sce_objects","clinical_data", "clinical_data_ROI_ac_combined_CORRECT.csv"))
clinical.data$TMA <-clinical.data$TMA.x
clinical.data$TMA.x <-NULL
clinical.data$TMA.y <-NULL
clinical.data$X.1 <-NULL
clinical.data$X <-NULL
head(clinical.data)

unique(clinical.data$DX.name)
clinical.data$DX.name[clinical.data$Patient_ID=="Control"] <-"Control"
table(clinical.data$DX.name)
clinical.data$Patient_ID %>% unique() %>% length() #1071 PATIENTS IN TOTAL

area <- read.csv(file=file.path(wd,"clinical_data", "area.csv"))
area$X <- NULL
area$Tma_ac <- area$TMA_ImageID
area$TMA_ImageID <-NULL
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

```

Table with Immune cell numbers including T cells removing undefined
```{r, Table with immune cell numbers per patient, echo=FALSE, message=FALSE, warning=FALSE}
#All T cells together
tbl <- as.data.frame(table(tcell.sce[,tcell.sce$Patient_ID!="Control" ]$Patient_ID))
colnames(tbl) <-c("Patient ID", "T cell number overall")
#write.csv(tbl, file=file.path(plot_folder, paste("Immune cell numbers per patient overall including T-excluding undefined cells.csv")))
#print(tbl)

#summary(tbl$`T cell number overall`) #3-6259
#tbl[tbl$`T cell number overall` <=100,] #31 patients have less than 100 T cells=3% but 10% of median/ 89 patients have less than 300 T cells

#Lowest 10% of all patients' T cell numbers
#tbl[order(tbl$`T cell number overall`),]$`T cell number overall`[1:100] #1-23


#Highest 10% of all patients' T cell numbers
#tbl[order(-tbl$`T cell number overall`),]$`T cell number overall`[1:100] #760-2378

#tbl[tbl$`T cell number overall` <=15,] #102 patients have less than 100 T cells=10%. 77 patients have less than 15 T cells

all_tcell_pat <- tbl[tbl$`T cell number overall` <=20,]$`Patient ID`

length(unique(tcell.sce$Patient_ID))
```

Remove ROIs
```{r, Table with Immune cell numbers per image, echo=FALSE, message=FALSE, warning=FALSE}
#All Immune cells together
tbl <- as.data.frame(table(tcell.sce[,tcell.sce$Patient_ID!="Control"]$RoiID))
colnames(tbl) <-c("ROI ID", "Immune cell number overall")
#write.csv(tbl, file=file.path(plot_folder, paste("Immune cell numbers per ROI overall including T cells-excluding undefined.csv")))
#print(tbl)
#summary(tbl$`Immune cell number overall`) #1-5075
#tbl[tbl$`Immune cell number overall` <=10,] #93 images have less than 50 Immune cells (10% of median)
#Lowest 10% of all patients' Immune cell numbers
#tbl[order(tbl$`Immune cell number overall`),]$`Immune cell number overall`[1:200] #1-110


#Highest 10% of all patients' Immune cell numbers
#tbl[order(-tbl$`Immune cell number overall`),]$`Immune cell number overall`[1:200] #1385-5075
#per image cut at 50 Immune cells per image equals lowest 5% -> ensures that there's at least 50 Immune cells per patient

all_tcell_roi <- tbl[tbl$`Immune cell number overall` <=10,]$`ROI ID`
```


Remove patients and roi
```{r, patient roi and pat_roi removal, echo=FALSE, message=FALSE, warning=FALSE}
#Patient removal
tcell.sce_pat <- tcell.sce[,tcell.sce$Patient_ID!="Control"&
                               !tcell.sce$Patient_ID%in%all_tcell_pat]
length(unique(tcell.sce_pat$Patient_ID)) #954

#Roi removal
tcell.sce_roi <- tcell.sce[,tcell.sce$Patient_ID!="Control"&
                               !tcell.sce$RoiID%in%all_tcell_roi]
length(unique(tcell.sce_roi$Patient_ID)) #987

#Patient & Roi removal
tcell.sce_pat.roi <- tcell.sce[,tcell.sce$Patient_ID!="Control"&
                               !tcell.sce$Patient_ID%in%all_tcell_pat&
                               !tcell.sce$RoiID%in%all_tcell_roi]
length(unique(tcell.sce_pat.roi$Patient_ID)) #954

```

```{r, Save SCE patient and roi removed, include=FALSE, echo=FALSE,warning=FALSE, message=FALSE, eval=FALSE}
data_folder <-(file.path(wd,"sce_objects","immune","Tcell"))

#saveRDS(tcell.sce_pat.roi,file=file.path(data_folder, paste("Tcell_sce_pat_roi_rem.rds",sep="")))
#saveRDS(tcell.sce_roi,file=file.path(data_folder, paste("Tcell_sce_roi_rem.rds",sep="")))
#saveRDS(tcell.sce_pat,file=file.path(data_folder, paste("Tcell_sce_pat_rem.rds",sep="")))

tcell.sce_pat.roi <-readRDS(file=file.path(data_folder, paste("Tcell_sce_pat_roi_rem.rds",sep="")))
tcell.sce_roi <-readRDS(file=file.path(data_folder, paste("Tcell_sce_roi_rem.rds",sep="")))
tcell.sce_pat <-readRDS(file=file.path(data_folder, paste("Tcell_sce_pat_rem.rds",sep="")))
```


##Analysis
#Proportions T cell CATEGORY

## **Analysis**
In the following, only tumours that were identified as adenocarcinomas or squamous cell carcinomas by pathologists will be used.

## **T cell Category**

## **Proportions**
Optimal number of patient metaclusters T cell category proportions
```{r, optimal number of clusters for non T cell category type proportion clusters, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Proportions
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
tdat <-as.data.frame(table( tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_type))


colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID
#tdat_m <-tdat_m[,1:2]
hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")

#devtools::install_github("kassambara/factoextra")
sbg <-cutree(hc, k=2)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=15)

#install.packages("cluster")
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
#res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

## Patients are ordered by hierarchical clustering (euclidian, wards.D2)
```{r, T cell category proportions ordered by hierarchical clustering, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table( tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = hc$labels[hc$order])
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = hc$labels[hc$order])

tdat_wide
```

Patient based metaclusters split by metacluster
```{r, Barplot T cell category proportions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#Barplot proportions ordered by
p <-ggplot(tdat,aes(x=freq, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()
#plot(p)
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_Tcell-Category_over_FILTEREDpatients_ordered-hc.pdf")), width=6, height=6)

#dendrogram coloured by metacluster
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 3) %>%  as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()+
  coord_flip()+
  scale_y_reverse()

#n
p2<-ggplot(tdat,aes(x=n, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+theme(
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())
plt <-p1+
  p+theme(legend.position = "none", axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())+
  p2

plot(plt)
#ggsave(plot=plt, file=file.path(plot_folder, paste("Barplot_n_hc_cell_type_proportions.pdf")), width=6, height=6)
```

```{r, merge hc metacluster into t_dat T cell category proportions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k= 3))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")

p <-ggplot(tdat_ct,aes(y=freq,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_split_metacluster_tcell-category_proportions.pdf")))
plot(p)
```


## **Survival analysis** - T cell category
T cell categories (CD4 and CD8) are grouped into "high" and "low" using the proportion's median as a cut-off.    
- high > median  
- low < median  
Patients who received neoadjuvant treatment are excluded from survival analysis.  
```{r, survival data T cell category proportions high-low, message=FALSE, warning=FALSE, echo=FALSE}
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat,]

#high low by median
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
surv <-c("CD8_G","CD4_G", "metacluster")
```


#CoxPH for T cell category proportions corrected for Stage, Grade and M
```{r, Coxph T cell category proportions corrected, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   CD4 + CD8 + Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `CD4+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r, survival T cell category proportions not split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}


#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  facet_wrap(~strata)+theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Split by tumour type (AC and SQC individually)
```{r, survival T cell category proportions split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#OS survival
surv <-c("CD8_G","CD4_G", "metacluster")

#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$CD4_G <- ifelse(dat$CD4 >summary(dat$CD4)[3],"CD4 high","CD4 low")
dat$CD8_G <- ifelse(dat$CD8 >summary(dat$CD8)[3],"CD8 high","CD8 low")
   
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
   #abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high / low as well as the proportion (continuous).   
```{r, coxph T cell category proportions, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("CD4_G","CD8_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("CD4_G","CD8_G","metacluster",colnames(clinical.data),contains("censor"))) #))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)

#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_Proportions_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Cell groups categorised into "high", "mid", "low".  
- high > 3rd quantile  
- mid >1st quantile and < 3rd quantile  
- low < 1st quantile  
```{r, survival T cell category proportions classified as high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv <-c("CD8_G","CD4_G", "metacluster")
```

## Survival T cell category proportions high mid low not split
```{r, survival t cell category proportions not split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Survival T cell category proportions high mid low split by tumour type
```{r, survival t cell category proportions split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$CD4_G[dat$CD4 >summary(dat$CD4)[5]] <-"CD4 high"
dat$CD4_G[dat$CD4 <summary(dat$CD4)[2]] <-"CD4 low"
dat$CD4_G[dat$CD4 >summary(dat$CD4)[2]& dat$CD4<summary(dat$CD4)[5]] <-"CD4 medium"

dat$CD8_G[dat$CD8 >summary(dat$CD8)[5]] <-"CD8 high"
dat$CD8_G[dat$CD8 <summary(dat$CD8)[2]] <-"CD8 low"
dat$CD8_G[dat$CD8 >summary(dat$CD8)[2]& dat$CD8<summary(dat$CD8)[5]] <-"CD8 medium"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high mid low as well as the proportion (continuous).   
```{r, coxph T cell category proportions high mid low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("CD4_G","CD8_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("CD4_G","CD8_G","metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_Proportions_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

#Correlation of T cell category proportions with clinical parameters
```{r, Correlation of non T cell category proportions and clinical parameters, message=FALSE, warning=FALSE, echo=FALSE}
tcell.sce_pat$NeoAdj <- ifelse(tcell.sce_pat$Chemo==1 |tcell.sce_pat$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                           tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")
tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

c.param <- c("Gender","T.new","N","M.new","Dist.Met","LN.Met","Stage")
for(i in c.param){
  print(i)
print(cor.test(tdat_wide_as$CD4,tdat_wide_as[[i]], method="spearman", exact=F))
  }
```


In group comparison (Wilcoxon for 2 groups or KW for >2 groups) for T cell categories over clinical variables:  
  - T.new (TNM)
  - N (TNM)
  - M.new (TNM)
  - Relapse
  - Grade
  - Stage
  - Gender
  - Typ (Tumour Type)
  - Dist.Met (Distant metastases yes/no)
  - LN.Met (Lymph node metastases yes/no)
  
## Kruskal-Wallis / Wilcoxon T cell category ~ clinical data excluding neoadjuvant therapy
```{r, KW W T cell category proportions WO neo, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]))

#remove neoadj patients
df <-df[!df$Patient_ID%in% neoadj_pat$Patient_ID,]
categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender") 

plot_list <- list()

for (i in (categories)) {

    t <- table(df$cell_type, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=2) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
  #pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_Proportions_WOneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```

## Neoadjuvantly treated patients are included, the same categories are used plus:  
- NeoAdj (Neoadjuvant therapy yes/no)
- Chemo (Neoadjuvant therapy = chemotherapy)
- Radio (Neoadjuvant therapy = radiotherapy)
```{r, KW W T cell category proportions with NEO, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]))

categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")

plot_list <- list()

for (i in (categories)) {

    t <- table(df$cell_type, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)
    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=2) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_Proportions_Wneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```



## **T cell category**
## **Densities**
Generally, densities are first calculated per tissue area. For patients with more than 1 image, the average density is calculated subsequently.  
Optimal number of T cell category densities patients metaclusters:
```{r, T cell category density optimal number of clusters, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Densities
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

#tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1]),"ward.D2")

#devtools::install_github("kassambara/factoextra")
library(FactoMineR)
library(factoextra)
sbg <-cutree(hc, k=8)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=20)

#install.packages("cluster")
library(cluster)
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

```{r, Calculate densities T cell category, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
  
  
#CAF type distribution over all patients
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID


hc <- hclust(dist(tdat_m[,-1]),"ward.D2")
#order Patients after clustering results
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = labels(hc))
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = labels(hc))
```

## Barplot showing absolute density per patietn coloured by T cell category together with hierarchical clustering tree coloured by patient metaclusters.
```{r,Barolot T cell category densities, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
#barolot (sqrt of density)
p <-ggplot(tdat,aes(y=sqrt(density),x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")

#barplot total density
p <-ggplot(tdat,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")


#coloured dendrogram
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 2) %>% as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()
pg <-plot_grid(p1,p, align="v", ncol=1)
pg
#ggsave(plot=pg, file=file.path(plot_folder, paste("Barplot_Tcell-Category-Densities_HC_ordered_dendro.pdf")), width=12, height=6)
```
Add patient metacluster to data and plot clusters separately for better identification
```{r,T cell category densities merge hc, message=FALSE, warning=FALSE, echo=FALSE}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k=2))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")
ggplot(tdat_ct,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
```

## Survival analysis over T cell category densities high low.  
- High > median
- Low < median
```{r, surv_dat T cell category densities high low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- surv_dat$CD4+surv_dat$CD8

#high low
surv_dat$Density_G <- ifelse(surv_dat$total_density >summary(surv_dat$total_density)[3],"Density high","Density low")
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
```

#CoxPH for T cell category density corrected for Stage, Grade and M
```{r, Coxph T cell category density corrected, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   CD4 + CD8 + Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `CD4+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r, survival T cell category densities high low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
#surv <-c("metacluster")
surv <-c("CD8_G","CD4_G","metacluster", "Density_G")
library(survminer)

for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
#km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## Split by tumour type (AC and SQC individually)
```{r, survival T cell category densities high low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Split
library(survminer)

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$Density_G <- ifelse(dat$total_density >summary(dat$total_density)[3],"Density high","Density low")
dat$CD4_G <- ifelse(dat$CD4 >summary(dat$CD4)[3],"CD4 high","CD4 low")
dat$CD8_G <- ifelse(dat$CD8 >summary(dat$CD8)[3],"CD8 high","CD8 low")   
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high / low as well as the densities (continuous).   
```{r, coxph T cell category densities high-low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=6, fig.width=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("Density_G","CD4_G","CD8_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("CD4_G","CD8_G","Density_G","metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_Density_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Survival T cell category densities high-mid-low     
- high    > 3rd quantile   
- mid     > 1st quantile < 3rd quantile   
- low     < 1st quantile   
```{r, surv_dat T cell category densities high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- surv_dat$CD4+surv_dat$CD8

#High Med Low
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[5]] <-"Density high"
surv_dat$Density_G[surv_dat$total_density <summary(surv_dat$total_density)[2]] <-"Density low"
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[2]& surv_dat$total_density<summary(surv_dat$total_density)[5]] <-"Density medium"

surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv <-c("CD4_G","CD8_G","Density_G","metacluster")
```

## T cell category densities split by tumour type
```{r, survival t cell category densities high-mid-low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
surv <-c("CD4_G","CD8_G","Density_G","metacluster")
library(survminer)

for (i in surv){
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## T cell category densities split by tumour type
```{r, survival t cell category densities high-mid-low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
surv <-c("CD4_G","CD8_G","Density_G","metacluster")

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$Density_G[dat$total_density >summary(dat$total_density)[5]] <-"Density high"
dat$Density_G[dat$total_density <summary(dat$total_density)[2]] <-"Density low"
dat$Density_G[dat$total_density >summary(dat$total_density)[2]& dat$total_density<summary(dat$total_density)[5]] <-"Density medium"

dat$CD4_G[dat$CD4 >summary(dat$CD4)[5]] <-"CD4 high"
dat$CD4_G[dat$CD4 <summary(dat$CD4)[2]] <-"CD4 low"
dat$CD4_G[dat$CD4 >summary(dat$CD4)[2]& dat$CD4<summary(dat$CD4)[5]] <-"CD4 medium"

dat$CD8_G[dat$CD8 >summary(dat$CD8)[5]] <-"CD8 high"
dat$CD8_G[dat$CD8 <summary(dat$CD8)[2]] <-"CD8 low"
dat$CD8_G[dat$CD8 >summary(dat$CD8)[2]& dat$CD8<summary(dat$CD8)[5]] <-"CD8 medium"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])


#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_type_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```



## Lasso-regressed cox-ph model   
including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high mid low as well as the densities (continuous).   
```{r,coxph T cell category densities high-mid-low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("Density_G","CD4_G","CD8_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("CD4_G","CD8_G","Density_G","metacluster",colnames(clinical.data),contains("censor")))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients

active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_Density_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))

#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables without neoadjuvant treated patients.
```{r, KW W T cell category density WO neo,warning=FALSE, message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$CD4 + tdat_wide$CD8
tdat <-tdat_wide %>% pivot_longer(cols=c("CD4","CD8","total_density"), names_to = "Phenotype",values_to = "density")

#remove
tdat <-tdat[!tdat$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)+
      theme(legend.position = "none")
    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=3) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
 # pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_Densities_WOneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables with neoadjuvant treated patients.
```{r, KW Wilcox T cell Category density W neo,warning=F,message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$CD4 + tdat_wide$CD8
tdat <-tdat_wide %>% pivot_longer(cols=c("CD4","CD8","total_density"), names_to = "Phenotype",values_to = "density")

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)+
      theme(legend.position = "none")
    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=3) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_Densities_Wneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

```{r, Correlation T cell category densities with clinical parameters, eval=F, echo=T, message=FALSE, warning=FALSE, echo=FALSE}
tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$CD4 + tdat_wide$CD8

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

for(i in c.param){
  print(i)
  print(cor.test(tdat_wide_as$CD4,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$CD8,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$total_density,tdat_wide_as[[i]], method="spearman", exact=F))
}
```


#T cell Type

##Analysis
#Proportions T cell CATEGORY

## **Analysis**
In the following, only tumours that were identified as adenocarcinomas or squamous cell carcinomas by pathologists will be used.

## **T cell Category**

## **DDensity**
Optimal number of patient metaclusters T cell category proportions
```{r, optimal number of clusters for non T cell category type density clusters, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Proportions
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
tdat <-as.data.frame(table( tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_subtype))


colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID
#tdat_m <-tdat_m[,1:2]
hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")

#devtools::install_github("kassambara/factoextra")
sbg <-cutree(hc, k=2)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=15)

#install.packages("cluster")
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
#res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

## Patients are ordered by hierarchical clustering (euclidian, wards.D2)
```{r, T cell category density ordered by hierarchical clustering, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table( tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            tcell.sce_pat[, tcell.sce_pat$Patient_ID!="Control"&
                                           tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                           tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = hc$labels[hc$order])
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = hc$labels[hc$order])

tdat_wide
```

Patient based metaclusters split by metacluster
```{r, Barplot T cell category density, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#Barplot density ordered by
p <-ggplot(tdat,aes(x=freq, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()
#plot(p)
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_Tcell-Category_over_FILTEREDpatients_ordered-hc.pdf")), width=6, height=6)

#dendrogram coloured by metacluster
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 8) %>%  as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()+
  coord_flip()+
  scale_y_reverse()

#n
p2<-ggplot(tdat,aes(x=n, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+theme(
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())

plt <-p1+
  p+theme(legend.position = "none", axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())+
  p2

plot(plt)
#ggsave(plot=plt, file=file.path(plot_folder, paste("Barplot_n_hc_cell_subtype_density.pdf")), width=6, height=6)
```

```{r, merge hc metacluster into t_dat T cell category density, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k= 8))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")

p <-ggplot(tdat_ct,aes(y=freq,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_split_metacluster_tcell-category_density.pdf")))
plot(p)
```


## **Survival analysis** - T cell category
T cell categories (CD4 and CD8) are grouped into "high" and "low" using the proportion's median as a cut-off.    
- high > median  
- low < median  
Patients who received neoadjuvant treatment are excluded from survival analysis.  
```{r, survival data T cell category density high-low, message=FALSE, warning=FALSE, echo=FALSE}
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat,]

#high low by median
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD4_Treg_G <- ifelse(surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[3],"CD4 Treg high","CD4 Treg low")
surv_dat$CD4_IDO_G <- ifelse(surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[3],"CD4 IDO high","CD4 IDO low")
surv_dat$CD4_PD1_G <- ifelse(surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[3],"CD4 PD1 high","CD4 PD1 low")
surv_dat$CD4_TCF_G <- ifelse(surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[3],"CD4 TCF1/7 high","CD4 TCF1/7 low")
surv_dat$CD4_ki67_G <- ifelse(surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[3],"dividing CD4 high","dividing CD4 low")

surv_dat$CD8_IDO_G <- ifelse(surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[3],"CD8 IDO high","CD8 IDO low")
surv_dat$CD8_TCF_G <- ifelse(surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[3],"CD8 TCF1/7 high","CD8 TCF1/7 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
surv_dat$CD8_ki67_G <- ifelse(surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[3],"dividing CD8 high","dividing CD8 low")

surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")
```


#CoxPH for T cell category density corrected for Stage, Grade and M
```{r, Coxph T cell category density, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   CD4 + CD8+`TCF1/7_CD4`+ki67_CD8+`TCF1/7_CD8`+PD1_CD4+IDO_CD8 +ki67_CD4+IDO_CD4+CD4_Treg+ Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `CD4+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r, survival T cell category density not split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}


#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  facet_wrap(~strata)+theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Split by tumour type (AC and SQC individually)
```{r, survival T cell category density split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#OS survival
surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster") 
#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]

#high low by median
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD4_Treg_G <- ifelse(surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[3],"CD4 Treg high","CD4 Treg low")
surv_dat$CD4_IDO_G <- ifelse(surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[3],"CD4 IDO high","CD4 IDO low")
surv_dat$CD4_PD1_G <- ifelse(surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[3],"CD4 PD1 high","CD4 PD1 low")
surv_dat$CD4_TCF_G <- ifelse(surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[3],"CD4 TCF1/7 high","CD4 TCF1/7 low")
surv_dat$CD4_ki67_G <- ifelse(surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[3],"dividing CD4 high","dividing CD4 low")

surv_dat$CD8_IDO_G <- ifelse(surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[3],"CD8 IDO high","CD8 IDO low")
surv_dat$CD8_TCF_G <- ifelse(surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[3],"CD8 TCF1/7 high","CD8 TCF1/7 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
surv_dat$CD8_ki67_G <- ifelse(surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[3],"dividing CD8 high","dividing CD8 low")


#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
   #abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high / low as well as the proportion (continuous).   
```{r, coxph T cell category density, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c(paste(surv),"Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c(paste(surv),colnames(clinical.data),contains("censor"))) #))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)

#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_density_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Cell groups categorised into "high", "mid", "low".  
- high > 3rd quantile  
- mid >1st quantile and < 3rd quantile  
- low < 1st quantile  
```{r, survival T cell category density classified as high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[5]] <-"CD4 Treg high"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg <summary(surv_dat$CD4_Treg)[2]] <-"CD4 Treg low"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[2]& surv_dat$CD4_Treg<summary(surv_dat$CD4_Treg)[5]] <-"CD4 T reg medium"

surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 high"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 <summary(surv_dat$IDO_CD4)[2]] <-"IDO_CD4 low"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[2]& surv_dat$IDO_CD4<summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 medium"

surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 high"
surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 <summary(surv_dat$PD1_CD4)[2]] <-"PD1_CD4 low"
surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[2]& surv_dat$PD1_CD4<summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 medium"

surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7 high"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` <summary(surv_dat$`TCF1/7_CD4`)[2]] <-"CD4 TCF1/7  low"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[2]& surv_dat$`TCF1/7_CD4`<summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7  medium"

surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing high"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 <summary(surv_dat$ki67_CD4)[2]] <-"CD4 dividing low"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[2]& surv_dat$ki67_CD4<summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO high"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 <summary(surv_dat$IDO_CD8)[2]] <-"CD8 IDO low"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[2]& surv_dat$IDO_CD8<summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO medium"

surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 high"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` <summary(surv_dat$`TCF1/7_CD8`)[2]] <-"CD8 TCF1/7 low"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[2]& surv_dat$`TCF1/7_CD8`<summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 medium"

surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing high"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 <summary(surv_dat$ki67_CD8)[2]] <-"CD8 dividing low"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[2]& surv_dat$ki67_CD8<summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing medium"

surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")
```

## Survival T cell category density high mid low not split
```{r, survival t cell category density not split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Survival T cell category density high mid low split by tumour type
```{r, survival t cell category density split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")

#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[5]] <-"CD4 Treg high"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg <summary(surv_dat$CD4_Treg)[2]] <-"CD4 Treg low"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[2]& surv_dat$CD4_Treg<summary(surv_dat$CD4_Treg)[5]] <-"CD4 T reg medium"

surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 high"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 <summary(surv_dat$IDO_CD4)[2]] <-"IDO_CD4 low"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[2]& surv_dat$IDO_CD4<summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 medium"

surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 high"
surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 <summary(surv_dat$PD1_CD4)[2]] <-"PD1_CD4 low"
surv_dat$PD1_CD4_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[2]& surv_dat$PD1_CD4<summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 medium"

surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7 high"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` <summary(surv_dat$`TCF1/7_CD4`)[2]] <-"CD4 TCF1/7  low"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[2]& surv_dat$`TCF1/7_CD4`<summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7  medium"

surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing high"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 <summary(surv_dat$ki67_CD4)[2]] <-"CD4 dividing low"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[2]& surv_dat$ki67_CD4<summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO high"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 <summary(surv_dat$IDO_CD8)[2]] <-"CD8 IDO low"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[2]& surv_dat$IDO_CD8<summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO medium"

surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 high"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` <summary(surv_dat$`TCF1/7_CD8`)[2]] <-"CD8 TCF1/7 low"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[2]& surv_dat$`TCF1/7_CD8`<summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 medium"

surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing high"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 <summary(surv_dat$ki67_CD8)[2]] <-"CD8 dividing low"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[2]& surv_dat$ki67_CD8<summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing medium"

#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high mid low as well as the proportion (continuous).   
```{r, coxph T cell category density high mid low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c(paste(surv),"Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c(paste(surv),"metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-Category_density_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

#Correlation of T cell category density with clinical parameters
```{r, Correlation of non T cell category density and clinical parameters, message=FALSE, warning=FALSE, echo=FALSE}
tcell.sce_pat$NeoAdj <- ifelse(tcell.sce_pat$Chemo==1 |tcell.sce_pat$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                           tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")
tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

c.param <- c("Gender","T.new","N","M.new","Dist.Met","LN.Met","Stage")
for(i in c.param){
  print(i)
print(cor.test(tdat_wide_as$CD4,tdat_wide_as[[i]], method="spearman", exact=F))
  }
```


In group comparison (Wilcoxon for 2 groups or KW for >2 groups) for T cell categories over clinical variables:  
  - T.new (TNM)
  - N (TNM)
  - M.new (TNM)
  - Relapse
  - Grade
  - Stage
  - Gender
  - Typ (Tumour Type)
  - Dist.Met (Distant metastases yes/no)
  - LN.Met (Lymph node metastases yes/no)
  
## Kruskal-Wallis / Wilcoxon T cell category ~ clinical data excluding neoadjuvant therapy
```{r, KW W T cell type density WO neo, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]))

#remove neoadj patients
df <-df[!df$Patient_ID%in% neoadj_pat$Patient_ID,]
categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender") 

plot_list <- list()

for (i in (categories)) {

    t <- table(df$cell_subtype, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)
    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=5) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
  #pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_density_WOneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```

## Neoadjuvantly treated patients are included, the same categories are used plus:  
- NeoAdj (Neoadjuvant therapy yes/no)
- Chemo (Neoadjuvant therapy = chemotherapy)
- Radio (Neoadjuvant therapy = radiotherapy)
```{r, KW W T cell category density with NEO, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(tcell.sce_pat[,tcell.sce_pat$Patient_ID!="Control"&
                                          tcell.sce_pat$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat$DX.name=="Squamous cell carcinoma"]))

categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")

plot_list <- list()

for (i in (categories)) {

    t <- table(df$cell_subtype, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)
    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=5) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-Category_density_Wneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```

Correlation of T cell category density per image
```{r, Correlations density per image T cell category, message=FALSE, warning=FALSE, echo=FALSE, eval=F}
tcell.sce_roi$NeoAdj <- ifelse(tcell.sce_roi$Chemo==1 |tcell.sce_roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_roi[,tcell.sce_roi$Patient_ID!="Control"&
                                          tcell.sce_roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_roi[,tcell.sce_roi$Patient_ID!="Control"&
                                          tcell.sce_roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$RoiID <-droplevels(tdat$RoiID)

tdat_wide <- pivot_wider(tdat,id_cols=c("RoiID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")
tdat_wide <-tdat_wide[!tdat_wide$RoiID%in% neoadj_pat$RoiID,]
rownames(tdat_wide)<-tdat_wide$RoiID


cor(tdat_wide[,-1], method="pearson")
#corrplot(cor(tdat_wide[,-1], method="pearson"), col=rev(brewer.pal(10,"RdBu")))

sig <- cor.mtest(tdat_wide[,-1], conf.level = .95) #is this test right?

corrplot(cor(tdat_wide[,-1], method="pearson"),
         tl.cex = .7,
         number.cex = .7,
         order = "hclust",
         p.mat = sig$p,
         insig = "label_sig",
         tl.col = "black",
         addrect = 3,
         tl.srt = 90,
         sig.level = c(.001, .01, .05),
         pch.cex = .9,
         col=rev(brewer.pal(10,"RdBu")),
         pch.col = "white",
         #type="upper",
         title = "Correlated ratios of CD4+ and CD8+ T cells",
        mar=c(0,0,3,0))
```


## **T cell Type**
## **Densities**
Generally, densities are first calculated per tissue area. For patients with more than 1 image, the average density is calculated subsequently.  
Optimal number of T cell density densities patients metaclusters:
```{r, T cell density density optimal number of clusters type, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Densities
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

#tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1]),"ward.D2")

#devtools::install_github("kassambara/factoextra")
library(FactoMineR)
library(factoextra)
sbg <-cutree(hc, k=8)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=20)

#install.packages("cluster")
library(cluster)
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

```{r, Calculate densities T cell type, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
  
  
#CAF type distribution over all patients
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID


hc <- hclust(dist(tdat_m[,-1]),"ward.D2")
#order Patients after clustering results
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = labels(hc))
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = labels(hc))
```

## Barplot showing absolute density per patietn coloured by T cell density together with hierarchical clustering tree coloured by patient metaclusters.
```{r,Barolot T cell type densities, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
#barolot (sqrt of density)
p <-ggplot(tdat,aes(y=sqrt(density),x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")

#barplot total density
p <-ggplot(tdat,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")


#coloured dendrogram
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 8) %>% as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()
pg <-plot_grid(p1,p, align="v", ncol=1)
pg
#ggsave(plot=pg, file=file.path(plot_folder, paste("Barplot_Tcell-density-Densities_HC_ordered_dendro.pdf")), width=12, height=6)
```
Add patient metacluster to data and plot clusters separately for better identification
```{r,T cell density densities merge hc, message=FALSE, warning=FALSE, echo=FALSE}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k=8))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")
ggplot(tdat_ct,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
```

## Survival analysis over T cell density densities high low.  
- High > median
- Low < median
```{r, surv_dat T cell density densities high low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- rowSums(surv_dat[,2:11])

#high low
surv_dat$Density_G <- ifelse(surv_dat$total_density >summary(surv_dat$total_density)[3],"Density high","Density low")
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD4_Treg_G <- ifelse(surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[3],"CD4 Treg high","CD4 Treg low")
surv_dat$CD4_IDO_G <- ifelse(surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[3],"CD4 IDO high","CD4 IDO low")
surv_dat$CD4_PD1_G <- ifelse(surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[3],"CD4 PD1 high","CD4 PD1 low")
surv_dat$CD4_TCF_G <- ifelse(surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[3],"CD4 TCF1/7 high","CD4 TCF1/7 low")
surv_dat$CD4_ki67_G <- ifelse(surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[3],"dividing CD4 high","dividing CD4 low")

surv_dat$CD8_IDO_G <- ifelse(surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[3],"CD8 IDO high","CD8 IDO low")
surv_dat$CD8_TCF_G <- ifelse(surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[3],"CD8 TCF1/7 high","CD8 TCF1/7 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
surv_dat$CD8_ki67_G <- ifelse(surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[3],"dividing CD8 high","dividing CD8 low")

surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")
```

#CoxPH for T cell density density corrected for Stage, Grade and M
```{r, Coxph T cell density density, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   CD4 + CD8+`TCF1/7_CD4`+ki67_CD8+`TCF1/7_CD8`+PD1_CD4+IDO_CD8 +ki67_CD4+IDO_CD4+CD4_Treg + Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `CD4+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r, survival T cell density densities high low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
#surv <-c("metacluster")
library(survminer)

for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
#km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## Split by tumour type (AC and SQC individually)
```{r, survival T cell density densities high low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")

#Split
library(survminer)

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
surv_dat$total_density <- rowSums(surv_dat[,2:11])
surv_dat$Density_G <- ifelse(surv_dat$total_density >summary(surv_dat$total_density)[3],"Density high","Density low")
surv_dat$CD4_G <- ifelse(surv_dat$CD4 >summary(surv_dat$CD4)[3],"CD4 high","CD4 low")
surv_dat$CD4_Treg_G <- ifelse(surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[3],"CD4 Treg high","CD4 Treg low")
surv_dat$CD4_IDO_G <- ifelse(surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[3],"CD4 IDO high","CD4 IDO low")
surv_dat$CD4_PD1_G <- ifelse(surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[3],"CD4 PD1 high","CD4 PD1 low")
surv_dat$CD4_TCF_G <- ifelse(surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[3],"CD4 TCF1/7 high","CD4 TCF1/7 low")
surv_dat$CD4_ki67_G <- ifelse(surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[3],"dividing CD4 high","dividing CD4 low")

surv_dat$CD8_IDO_G <- ifelse(surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[3],"CD8 IDO high","CD8 IDO low")
surv_dat$CD8_TCF_G <- ifelse(surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[3],"CD8 TCF1/7 high","CD8 TCF1/7 low")
surv_dat$CD8_G <- ifelse(surv_dat$CD8 >summary(surv_dat$CD8)[3],"CD8 high","CD8 low")
surv_dat$CD8_ki67_G <- ifelse(surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[3],"dividing CD8 high","dividing CD8 low")

#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high / low as well as the densities (continuous).   
```{r, coxph T cell density densities high-low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=6, fig.width=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c(paste(surv),"Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c(paste(surv),"Density_G","metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-density_Density_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Survival T cell density densities high-mid-low     
- high    > 3rd quantile   
- mid     > 1st quantile < 3rd quantile   
- low     < 1st quantile   
```{r, surv_dat T cell density densities high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- rowSums(surv_dat[,2:11])

#High Med Low
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[5]] <-"Density high"
surv_dat$Density_G[surv_dat$total_density <=summary(surv_dat$total_density)[2]] <-"Density low"
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[2]& surv_dat$total_density<summary(surv_dat$total_density)[5]] <-"Density medium"

surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <=summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[5]] <-"CD4 Treg high"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg <=summary(surv_dat$CD4_Treg)[2]] <-"CD4 Treg low"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[2]& surv_dat$CD4_Treg<summary(surv_dat$CD4_Treg)[5]] <-"CD4 T reg medium"

surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 high"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 <=summary(surv_dat$IDO_CD4)[2]] <-"IDO_CD4 low"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[2]& surv_dat$IDO_CD4<summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 medium"

surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 high"
surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 <=summary(surv_dat$PD1_CD4)[2]] <-"PD1_CD4 low"
surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[2]& surv_dat$PD1_CD4<summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 medium"

surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7 high"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` <=summary(surv_dat$`TCF1/7_CD4`)[2]] <-"CD4 TCF1/7  low"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[2]& surv_dat$`TCF1/7_CD4`<summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7  medium"

surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing high"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 <=summary(surv_dat$ki67_CD4)[2]] <-"CD4 dividing low"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[2]& surv_dat$ki67_CD4<summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <=summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO high"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 <=summary(surv_dat$IDO_CD8)[2]] <-"CD8 IDO low"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[2]& surv_dat$IDO_CD8<summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO medium"

surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 high"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` <=summary(surv_dat$`TCF1/7_CD8`)[2]] <-"CD8 TCF1/7 low"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[2]& surv_dat$`TCF1/7_CD8`<summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 medium"

surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing high"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 <=summary(surv_dat$ki67_CD8)[2]] <-"CD8 dividing low"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[2]& surv_dat$ki67_CD8<summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing medium"

surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")
```

## T cell density densities split by tumour type
```{r, survival t cell density densities high-mid-low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
library(survminer)

for (i in surv){
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## T cell density densities split by tumour type
```{r, survival t cell density densities high-mid-low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
surv <-c("CD8_G","CD4_G","CD4_Treg_G","CD4_IDO_G","CD4_PD1_G","CD4_TCF_G","CD4_ki67_G", "CD8_IDO_G","CD8_TCF_G","CD8_ki67_G","metacluster")

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[5]] <-"Density high"
surv_dat$Density_G[surv_dat$total_density <=summary(surv_dat$total_density)[2]] <-"Density low"
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[2]& surv_dat$total_density<summary(surv_dat$total_density)[5]] <-"Density medium"

surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[5]] <-"CD4 high"
surv_dat$CD4_G[surv_dat$CD4 <=summary(surv_dat$CD4)[2]] <-"CD4 low"
surv_dat$CD4_G[surv_dat$CD4 >summary(surv_dat$CD4)[2]& surv_dat$CD4<summary(surv_dat$CD4)[5]] <-"CD4 medium"

surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[5]] <-"CD4 Treg high"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg <=summary(surv_dat$CD4_Treg)[2]] <-"CD4 Treg low"
surv_dat$CD4_Treg_G[surv_dat$CD4_Treg >summary(surv_dat$CD4_Treg)[2]& surv_dat$CD4_Treg<summary(surv_dat$CD4_Treg)[5]] <-"CD4 T reg medium"

surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 high"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 <=summary(surv_dat$IDO_CD4)[2]] <-"IDO_CD4 low"
surv_dat$CD4_IDO_G[surv_dat$IDO_CD4 >summary(surv_dat$IDO_CD4)[2]& surv_dat$IDO_CD4<summary(surv_dat$IDO_CD4)[5]] <-"IDO_CD4 medium"

surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 high"
surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 <=summary(surv_dat$PD1_CD4)[2]] <-"PD1_CD4 low"
surv_dat$CD4_PD1_G[surv_dat$PD1_CD4 >summary(surv_dat$PD1_CD4)[2]& surv_dat$PD1_CD4<summary(surv_dat$PD1_CD4)[5]] <-"PD1_CD4 medium"

surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7 high"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` <=summary(surv_dat$`TCF1/7_CD4`)[2]] <-"CD4 TCF1/7  low"
surv_dat$CD4_TCF_G[surv_dat$`TCF1/7_CD4` >summary(surv_dat$`TCF1/7_CD4`)[2]& surv_dat$`TCF1/7_CD4`<summary(surv_dat$`TCF1/7_CD4`)[5]] <-"CD4 TCF1/7  medium"

surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing high"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 <=summary(surv_dat$ki67_CD4)[2]] <-"CD4 dividing low"
surv_dat$CD4_ki67_G[surv_dat$ki67_CD4 >summary(surv_dat$ki67_CD4)[2]& surv_dat$ki67_CD4<summary(surv_dat$ki67_CD4)[5]] <-"CD4 dividing medium"

surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[5]] <-"CD8 high"
surv_dat$CD8_G[surv_dat$CD8 <=summary(surv_dat$CD8)[2]] <-"CD8 low"
surv_dat$CD8_G[surv_dat$CD8 >summary(surv_dat$CD8)[2]& surv_dat$CD8<summary(surv_dat$CD8)[5]] <-"CD8 medium"

surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO high"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 <=summary(surv_dat$IDO_CD8)[2]] <-"CD8 IDO low"
surv_dat$CD8_IDO_G[surv_dat$IDO_CD8 >summary(surv_dat$IDO_CD8)[2]& surv_dat$IDO_CD8<summary(surv_dat$IDO_CD8)[5]] <-"CD8 IDO medium"

surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 high"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` <=summary(surv_dat$`TCF1/7_CD8`)[2]] <-"CD8 TCF1/7 low"
surv_dat$CD8_TCF_G[surv_dat$`TCF1/7_CD8` >summary(surv_dat$`TCF1/7_CD8`)[2]& surv_dat$`TCF1/7_CD8`<summary(surv_dat$`TCF1/7_CD8`)[5]] <-"CD8 TCF1/7 medium"

surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing high"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 <=summary(surv_dat$ki67_CD8)[2]] <-"CD8 dividing low"
surv_dat$CD8_ki67_G[surv_dat$ki67_CD8 >summary(surv_dat$ki67_CD8)[2]& surv_dat$ki67_CD8<summary(surv_dat$ki67_CD8)[5]] <-"CD8 dividing medium"

#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])


#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-cell_subtype_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```



## Lasso-regressed cox-ph model   
including clinical parameters, patient metacluster and T cell categories (CD4 vs CD8) classified as high mid low as well as the densities (continuous).   
```{r,coxph T cell density densities high-mid-low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c(paste(surv),"Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c(paste(surv),"metacluster",colnames(clinical.data),contains("censor")))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients

active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tcell-density_Density_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))

#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables without neoadjuvant treated patients.
```{r, KW W T cell density density WO neo,warning=FALSE, message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- rowSums(tdat_wide[,2:11])
tdat <-tdat_wide %>% pivot_longer(cols=c(colnames(tdat_wide[,2:12])), names_to = "Phenotype",values_to = "density")

#remove
tdat <-tdat[!tdat$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)
    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.ticks.x = element_blank(),
              legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=6) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
 # pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-density_Densities_WOneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables with neoadjuvant treated patients.
```{r, KW Wilcox T cell density density W neo,warning=FALSE, message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- rowSums(tdat_wide[,2:11])
tdat <-tdat_wide %>% pivot_longer(cols=c(colnames(tdat_wide[,2:12])), names_to = "Phenotype",values_to = "density")

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.ticks.x = element_blank(),
              legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=6) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tcell-density_Densities_Wneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

```{r, Correlation T cell density densities with clinical parameters, eval=F, echo=T, message=FALSE, warning=FALSE, echo=FALSE}
tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$CD4 + tdat_wide$CD8

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

for(i in c.param){
  print(i)
  print(cor.test(tdat_wide_as$CD4,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$CD8,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$total_density,tdat_wide_as[[i]], method="spearman", exact=F))
}
```



## Correlation of T cell type densites amongst eachother
```{r, Correlations T cell type densities amongst eachother, message=FALSE, warning=FALSE, echo=FALSE, fig.width=6, fig.height=6}
tcell.sce_pat.roi$NeoAdj <- ifelse(tcell.sce_pat.roi$Chemo==1 |tcell.sce_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           tcell.sce_pat.roi[,tcell.sce_pat.roi$Patient_ID!="Control"&
                                          tcell.sce_pat.roi$DX.name=="Adenocarcinoma"|
                                          tcell.sce_pat.roi$DX.name=="Squamous cell carcinoma"]$cell_subtype))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=tcell.sce_pat.roi$RoiID,
                                "Area"=tcell.sce_pat.roi$Area_px_Core,
                                "Patient_ID"=tcell.sce_pat.roi$Patient_ID,
                                "DX.name"=tcell.sce_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

tdat <-tdat %>% 
  group_by(RoiID,Phenotype) %>%
  summarise(across(-c(Patient_ID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
tdat_wide <- pivot_wider(tdat,id_cols=c("RoiID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

neoadj_roi <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(RoiID) 

tdat_wide <-tdat_wide[!tdat_wide$RoiID%in% neoadj_roi$RoiID,]
rownames(tdat_wide)<-tdat_wide$RoiID

library(corrplot)
cor(tdat_wide[,-1], method="pearson")
#corrplot(cor(tdat_wide[,-1], method="pearson"), col=rev(brewer.pal(10,"RdBu")))

#test <- immune.pheno.sub %>% select(contains("Ratio"))
#test$Ratio_tCAF_vCAF <-NULL
sig <- cor.mtest(tdat_wide[,-1], conf.level = .95) #is this test right?


#plot
corrplot(cor(tdat_wide[,-1], method="pearson"),
         tl.cex = .7,
         number.cex = .7,
         order = "hclust",
         p.mat = sig$p,
         insig = "label_sig",
         tl.col = "black",
         addrect = 3,
         tl.srt = 90,
         sig.level = c(.001, .01, .05),
         pch.cex = .9,
         col=rev(brewer.pal(10,"RdBu")),
         pch.col = "white",
         #type="upper",
         title = "Correlation of T cell type densities",
        mar=c(0,0,3,0))
```
