---
title: "R Notebook - Analysis Tumour cells"
output:
  html_document:
    df_print: paged
---

```{r  import libraries, echo=F, warnings=F, message=F}
library(igraph)
library(SingleCellExperiment)
library(S4Vectors)
library(stringr)
library(DT)
library(dplyr)
library(tidyr)
library(mclust)
library(ggplot2)
library(RColorBrewer)
library(scater)
library(Rphenoannoy)
library(factoextra)
library(cluster)
library(dendextend)
library(ggthemes)
library(ggpubr)
library(dplyr)
#library(tidyverse)
library(RColorBrewer)
library(pals)
library(qwraps2)
library(table1)
library(SingleCellExperiment)
library(tidyr)
library(scater)
library(ggridges)
library(viridis)
library(viridisLite)
library(ggplot2)
library(data.table)
library(CATALYST)
library(gridExtra)
library(Rphenograph)
library(ComplexHeatmap)
library(CATALYST)
library(scales)
library(survival)
library(broom)
library(pheatmap)
library(FlowSOM)
library(Seurat)
library(Rphenoannoy)
library(dplyr)
library(data.table)
library(ggthemes)
library(diffcyt)
library(edgeR)
library(rstatix)
library(dendextend)
library(ggdendro)
library(dendextend)
library(FactoMineR)
library(factoextra)
library(survminer)
library(corrplot)
library(rstatix)
library(graphics)
library(cowplot)
library(cluster)
library(glmnet)
library(fastDummies)
library(ggsci)
library(FactoMineR)
library(factoextra)
set.seed(101100)
```



```{r  Set wd and load data final vessel, message=F, warning=F, echo=F}
#set working directory 
wd <-dirname(getwd())

data_folder <-(file.path(wd,"sce_objects","final objects with categories","FINAL"))

#RAW
#all.tumour <-readRDS(file=file.path(data_folder, "TUMOUR_CLINICAL-DATA_FILTERED.rds"))

data_folder <-(file.path(wd,"sce_objects","Tumour"))

#saveRDS(all.tumour, file=file.path(data_folder, "FINAL_All_Tumour_clustered.rds"))

#workingfile
all.tumour <- readRDS(file=file.path(data_folder, "FINAL_All_Tumour_clustered.rds"))

all.tumour$DX.name[is.na(all.tumour$DX.name)]<-"NA"

```

Tumour marker
```{r  Define final tumour markers, echo=F, warning=F, message=FALSE}
all.marker <-rownames(all.tumour)

bad.marker <- c("Iridium_191","Iridium_193","Cadherin-6","Histone H3") 
good.marker <- all.marker[!all.marker %in% bad.marker]

tumour.marker <-c("Carbonic Anhydrase IX","Pan Cytokeratin + Keratin Epithelial")
print(tumour.marker)
```

```{r  load clinical data, echo=F, warnings=F, message=F}
clinical.data <- read.csv(file=file.path(wd,"sce_objects","clinical_data", "clinical_data_ROI_ac_combined_CORRECT.csv"))
clinical.data$TMA <-clinical.data$TMA.x
clinical.data$TMA.x <-NULL
clinical.data$TMA.y <-NULL
clinical.data$X.1 <-NULL
clinical.data$X <-NULL
head(clinical.data)

unique(clinical.data$DX.name)
clinical.data$DX.name[clinical.data$Patient_ID=="Control"] <-"Control"
table(clinical.data$DX.name)
clinical.data$Patient_ID %>% unique() %>% length() #1071 PATIENTS IN TOTAL

area <- read.csv(file=file.path(wd,"clinical_data", "area.csv"))
area$X <- NULL
area$Tma_ac <- area$TMA_ImageID
area$TMA_ImageID <-NULL
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

```

Table with Immune cell numbers including tumour removing undefined
```{r  Table with tumour cell numbers per patient, echo=FALSE, message=FALSE, warning=FALSE}
#All tumour together
tbl <- as.data.frame(table(all.tumour[,all.tumour$Patient_ID!="Control" ]$Patient_ID))
colnames(tbl) <-c("Patient ID", "Tumour cell number overall")
#write.csv(tbl, file=file.path(plot_folder, paste("Immune cell numbers per patient overall including T-excluding undefined cells.csv")))
#print(tbl)

#summary(tbl$`Tumour cell number overall`) #3-6259
#tbl[tbl$`Tumour cell number overall` <=100,] #31 patients have less than 100 tumour=3% but 10% of median/ 89 patients have less than 300 tumour

#Lowest 10% of all patients' Tumour cell numbers
#tbl[order(tbl$`Tumour cell number overall`),]$`Tumour cell number overall`[1:100] #2-502


#Highest 10% of all patients' Tumour cell numbers
#tbl[order(-tbl$`Tumour cell number overall`),]$`Tumour cell number overall`[1:100] #5133-9247

#tbl[tbl$`Tumour cell number overall` <=500,] #102 patients have less than 100 tumour=10%. 77 patients have less than 15 tumour

all_Tumour_pat <- tbl[tbl$`Tumour cell number overall` <=50,]$`Patient ID`

length(unique(all.tumour$Patient_ID))
```

Remove ROIs
```{r  Table with tumour cell numbers per image, echo=FALSE, message=FALSE, warning=FALSE}
#All Immune cells together
tbl <- as.data.frame(table(all.tumour[,all.tumour$Patient_ID!="Control"]$RoiID))
colnames(tbl) <-c("ROI ID", "Immune cell number overall")
#write.csv(tbl, file=file.path(plot_folder, paste("Immune cell numbers per ROI overall including tumour-excluding undefined.csv")))
#print(tbl)
#summary(tbl$`Immune cell number overall`) #1-5075
#tbl[tbl$`Immune cell number overall` <=200,] #93 images have less than 50 Immune cells (10% of median)
#Lowest 10% of all patients' Immune cell numbers
#tbl[order(tbl$`Immune cell number overall`),]$`Immune cell number overall`[1:200] #1-195


#Highest 10% of all patients' Immune cell numbers
#tbl[order(-tbl$`Immune cell number overall`),]$`Immune cell number overall`[1:200] #1385-5075
#per image cut at 50 Immune cells per image equals lowest 5% -> ensures that there's at least 50 Immune cells per patient

all_Tumour_roi <- tbl[tbl$`Immune cell number overall` <=200,]$`ROI ID`
```


Remove patients and roi
```{r  patient roi and pat_roi removal, echo=FALSE, message=FALSE, warning=FALSE}
#Patient removal
all.tumour_pat <- all.tumour[,all.tumour$Patient_ID!="Control"&
                               !all.tumour$Patient_ID%in%all_Tumour_pat]
length(unique(all.tumour_pat$Patient_ID)) #954

#Roi removal
all.tumour_roi <- all.tumour[,all.tumour$Patient_ID!="Control"&
                               !all.tumour$RoiID%in%all_Tumour_roi]
length(unique(all.tumour_roi$Patient_ID)) #987

#Patient & Roi removal
all.tumour_pat.roi <- all.tumour[,all.tumour$Patient_ID!="Control"&
                               !all.tumour$Patient_ID%in%all_Tumour_pat&
                               !all.tumour$RoiID%in%all_Tumour_roi]
length(unique(all.tumour_pat.roi$Patient_ID)) #954

```

```{r  Save SCE patient and roi removed, include=FALSE, echo=FALSE,warning=FALSE, message=FALSE, eval=FALSE}
data_folder <-(file.path(wd,"sce_objects","Tumour"))

all.tumour_pat.roi$DX.name[is.na(all.tumour_pat.roi$DX.name)]<-"NA"
all.tumour_roi$DX.name[is.na(all.tumour_roi$DX.name)]<-"NA"
all.tumour_pat$DX.name[is.na(all.tumour_pat$DX.name)]<-"NA"

saveRDS(all.tumour_pat.roi,file=file.path(data_folder, paste("Tumour_sce_pat_roi_rem.rds",sep="")))
saveRDS(all.tumour_roi,file=file.path(data_folder, paste("Tumour_sce_roi_rem.rds",sep="")))
saveRDS(all.tumour_pat,file=file.path(data_folder, paste("Tumour_sce_pat_rem.rds",sep="")))

all.tumour_pat.roi <-readRDS(file=file.path(data_folder, paste("Tumour_sce_pat_roi_rem.rds",sep="")))
all.tumour_roi <-readRDS(file=file.path(data_folder, paste("Tumour_sce_roi_rem.rds",sep="")))
all.tumour_pat <-readRDS(file=file.path(data_folder, paste("Tumour_sce_pat_rem.rds",sep="")))


```


##Analysis
#Proportions tumour type

## **Analysis**
In the following, only tumours that were identified as adenocarcinomas or squamous cell carcinomas by pathologists will be used.

## **tumour type**

## **Proportions**
Optimal number of patient metaclusters tumour type proportions
```{r  optimal number of clusters for tumour type proportion clusters, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Proportions
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
tdat <-as.data.frame(table( all.tumour_pat[, all.tumour_pat$Patient_ID!="Control"&
                                           all.tumour_pat$DX.name=="Adenocarcinoma"|
                                           all.tumour_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            all.tumour_pat[, all.tumour_pat$Patient_ID!="Control"&
                                           all.tumour_pat$DX.name=="Adenocarcinoma"|
                                           all.tumour_pat$DX.name=="Squamous cell carcinoma"]$tumour_type))


colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID
#tdat_m <-tdat_m[,1:2]
hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")

#devtools::install_github("kassambara/factoextra")
sbg <-cutree(hc, k=2)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=15)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=15)

#install.packages("cluster")
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
#res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

## Patients are ordered by hierarchical clustering (euclidian, wards.D2)
```{r  tumour type proportions ordered by hierarchical clustering, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table( all.tumour_pat[, all.tumour_pat$Patient_ID!="Control"&
                                           all.tumour_pat$DX.name=="Adenocarcinoma"|
                                           all.tumour_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                            all.tumour_pat[, all.tumour_pat$Patient_ID!="Control"&
                                           all.tumour_pat$DX.name=="Adenocarcinoma"|
                                           all.tumour_pat$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1], method="euclidian"),"ward.D2")
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = hc$labels[hc$order])
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = hc$labels[hc$order])

tdat_wide
```

Patient based metaclusters split by metacluster
```{r  Barplot tumour type proportions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#Barplot proportions ordered by
p <-ggplot(tdat,aes(x=freq, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()
#plot(p)
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_Tumour-Category_over_FILTEREDpatients_ordered-hc.pdf")), width=6, height=6)

#dendrogram coloured by metacluster
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 4) %>%  as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()+
  coord_flip()+
  scale_y_reverse()

#n
p2<-ggplot(tdat,aes(x=n, y=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+theme(
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())
plt <-p1+
  p+theme(legend.position = "none", axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank())+
  p2

plot(plt)
#ggsave(plot=plt, file=file.path(plot_folder, paste("Barplot_n_hc_tumour_type_proportions.pdf")), width=6, height=6)
```

```{r  merge hc metacluster into t_dat tumour type proportions, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k= 4))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")

p <-ggplot(tdat_ct,aes(y=freq,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
#ggsave(plot=p, file=file.path(plot_folder, paste("Barplot_split_metacluster_Tumour-category_proportions.pdf")))
plot(p)
```


## **Survival analysis** - tumour type
T cell categories (normal and hypoxic) are grouped into "high" and "low" using the proportion's median as a cut-off.    
- high > median  
- low < median  
Patients who received neoadjuvant treatment are excluded from survival analysis.  
```{r  survival data tumour type proportions high-low, message=FALSE, warning=FALSE, echo=FALSE}
neoadj_pat <- clinical.data %>% filter(NeoAdj == "NeoAdjuvantTherapy") %>% select(Patient_ID) 
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat,]

#high low by median
surv_dat$normal_G <- ifelse(surv_dat$normal >summary(surv_dat$normal)[3],"normal high","normal low")
surv_dat$hypoxic_G <- ifelse(surv_dat$hypoxic >summary(surv_dat$hypoxic)[3],"hypoxic high","hypoxic low")
surv <-c("hypoxic_G","normal_G", "metacluster")
```


#CoxPH for tumour type proportions corrected for Stage, Grade and M
```{r  Coxph tumour type proportions corrected, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   normal + hypoxic + Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `normal+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r  survival tumour type proportions not split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}


#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  facet_wrap(~strata)+theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Split by tumour type (AC and SQC individually)
```{r  survival tumour type proportions split high low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#OS survival
surv <-c("hypoxic_G","normal_G", "metacluster")

#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$normal_G <- ifelse(dat$normal >summary(dat$normal)[3],"normal high","normal low")
dat$hypoxic_G <- ifelse(dat$hypoxic >summary(dat$hypoxic)[3],"hypoxic high","hypoxic low")
   
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
   #abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (normal vs hypoxic) classified as high / low as well as the proportion (continuous).   
```{r  coxph tumour type proportions, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("normal_G","hypoxic_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("normal_G","hypoxic_G","metacluster",colnames(clinical.data),contains("censor"))) #))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)

#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tumour-Category_Proportions_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Cell groups categorised into "high", "mid", "low".  
- high > 3rd quantile  
- mid >1st quantile and < 3rd quantile  
- low < 1st quantile  
```{r  survival tumour type proportions classified as high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat$normal_G[surv_dat$normal >summary(surv_dat$normal)[5]] <-"normal high"
surv_dat$normal_G[surv_dat$normal <summary(surv_dat$normal)[2]] <-"normal low"
surv_dat$normal_G[surv_dat$normal >summary(surv_dat$normal)[2]& surv_dat$normal<summary(surv_dat$normal)[5]] <-"normal medium"

surv_dat$hypoxic_G[surv_dat$hypoxic >summary(surv_dat$hypoxic)[5]] <-"hypoxic high"
surv_dat$hypoxic_G[surv_dat$hypoxic <summary(surv_dat$hypoxic)[2]] <-"hypoxic low"
surv_dat$hypoxic_G[surv_dat$hypoxic >summary(surv_dat$hypoxic)[2]& surv_dat$hypoxic<summary(surv_dat$hypoxic)[5]] <-"hypoxic medium"

surv <-c("hypoxic_G","normal_G", "metacluster")
```

## Survival tumour type proportions high mid low not split
```{r  survival tumour type proportions not split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#not split by Tumour Type (AC/SQC)
for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
  # symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
```

## Survival tumour type proportions high mid low split by tumour type
```{r  survival tumour type proportions split high mid low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#split by Tumour Type (AC/SQC)
for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$normal_G[dat$normal >summary(dat$normal)[5]] <-"normal high"
dat$normal_G[dat$normal <summary(dat$normal)[2]] <-"normal low"
dat$normal_G[dat$normal >summary(dat$normal)[2]& dat$normal<summary(dat$normal)[5]] <-"normal medium"

dat$hypoxic_G[dat$hypoxic >summary(dat$hypoxic)[5]] <-"hypoxic high"
dat$hypoxic_G[dat$hypoxic <summary(dat$hypoxic)[2]] <-"hypoxic low"
dat$hypoxic_G[dat$hypoxic >summary(dat$hypoxic)[2]& dat$hypoxic<summary(dat$hypoxic)[5]] <-"hypoxic medium"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = surv_dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
 #  symbols = c("****", "***", "**", "*", "+", "."),
  # abbr.colnames = FALSE, na = " "))
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)   
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,2000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
including clinical parameters, patient metacluster and T cell categories (normal vs hypoxic) classified as high mid low as well as the proportion (continuous).   
```{r  coxph tumour type proportions high mid low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("normal_G","hypoxic_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("normal_G","hypoxic_G","metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tumour-Category_Proportions_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

#Correlation of tumour type proportions with clinical parameters
```{r  Correlation of tumour type proportions and clinical parameters, message=FALSE, warning=FALSE, echo=FALSE}
all.tumour_pat$NeoAdj <- ifelse(all.tumour_pat$Chemo==1 |all.tumour_pat$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(all.tumour_pat[,all.tumour_pat$Patient_ID!="Control"&
                                          all.tumour_pat$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat$DX.name=="Squamous cell carcinoma"]$Patient_ID,
                           all.tumour_pat[,all.tumour_pat$Patient_ID!="Control"&
                                          all.tumour_pat$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("Patient_ID","Phenotype","n")

tdat <-tdat %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::mutate(freq = n / sum(n))
tdat[is.na(tdat)] <- 0
tdat$Phenotype <-droplevels(tdat$Phenotype)
tdat$Patient_ID <-droplevels(tdat$Patient_ID)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="freq",names_prefix = "")
tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

c.param <- c("Gender","T.new","N","M.new","Dist.Met","LN.Met","Stage")
for(i in c.param){
  print(i)
print(cor.test(tdat_wide_as$normal,tdat_wide_as[[i]], method="spearman", exact=F))
  }
```


In group comparison (Wilcoxon for 2 groups or KW for >2 groups) for T cell categories over clinical variables:  
  - T.new (TNM)
  - N (TNM)
  - M.new (TNM)
  - Relapse
  - Grade
  - Stage
  - Gender
  - Typ (Tumour Type)
  - Dist.Met (Distant metastases yes/no)
  - LN.Met (Lymph node metastases yes/no)
  
## Kruskal-Wallis / Wilcoxon tumour type ~ clinical data excluding neoadjuvant therapy
```{r  KW W tumour type proportions WO neo, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(all.tumour_pat[,all.tumour_pat$Patient_ID!="Control"&
                                          all.tumour_pat$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat$DX.name=="Squamous cell carcinoma"]))

#remove neoadj patients
df <-df[!df$Patient_ID%in% neoadj_pat$Patient_ID,]
categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender") 

plot_list <- list()

for (i in (categories)) {

    t <- table(df$tumour_type, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=2) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
  #pdf(file=file.path(plot_folder, paste0("KW_W_Tumour-Category_Proportions_WOneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```

## Neoadjuvantly treated patients are included, the same categories are used plus:  
- NeoAdj (Neoadjuvant therapy yes/no)
- Chemo (Neoadjuvant therapy = chemotherapy)
- Radio (Neoadjuvant therapy = radiotherapy)
```{r  KW W tumour type proportions with NEO, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
df <- as.data.frame(colData(all.tumour_pat[,all.tumour_pat$Patient_ID!="Control"&
                                          all.tumour_pat$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat$DX.name=="Squamous cell carcinoma"]))

categories <- c("Dist.Met","LN.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")

plot_list <- list()

for (i in (categories)) {

    t <- table(df$tumour_type, df$Patient_ID)
    t <- prop.table(t, margin=2)
    t <- as.data.frame(t)
    colnames(t) <- c("Phenotype", "Patient_ID","freq")
    t<-merge(t, clinical.data, by.x="Patient_ID", by.y="Patient_ID")
    tdat <-t
    tdat$i <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$Phenotype <-droplevels(tdat$Phenotype)
  
  for (j in unique(t$Phenotype)){
    tdat<-t
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(freq~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
      geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=4, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Proportion")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)
    }
    
    else {
if(is.element(i,colnames(clinical.data)) == TRUE){
    
      tdat<-t
      tdat[[i]] <-as.factor(tdat[[i]])
      tdat <-tdat %>% drop_na(i)
      tdat$Phenotype <-droplevels(tdat$Phenotype)

      tdat <-subset(tdat, Phenotype==j)
      detach("package:dplyr", unload=TRUE)
      library(dplyr)
      
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$freq~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))

      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = freq, colour=.data[[i]]))+
        geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=4, strip.position="top")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(x=paste(i), y="Proportion", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }
   }  

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=2) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tumour-Category_Proportions_Wneo",i,".pdf")), width=10, height=5)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=length(unique(t$Phenotype)))
 # dev.off()
}
```



## **tumour type**
## **Densities**
Generally, densities are first calculated per tissue area. For patients with more than 1 image, the average density is calculated subsequently.  
Optimal number of tumour type densities patients metaclusters:
```{r  tumour type density optimal number of clusters, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=4}
#Calculate Densities
tdat <-as.data.frame(table(all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=all.tumour_pat.roi$RoiID,
                                "Area"=all.tumour_pat.roi$Area_px_Core,
                                "Patient_ID"=all.tumour_pat.roi$Patient_ID,
                                "DX.name"=all.tumour_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

#tdat_wide <-tdat_wide[!tdat_wide$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID

hc <- hclust(dist(tdat_m[,-1]),"ward.D2")

#devtools::install_github("kassambara/factoextra")
library(FactoMineR)
library(factoextra)
sbg <-cutree(hc, k=8)
fviz_cluster(list(data = tdat_wide[,-1], cluster = sbg))

fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "wss", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "gap_stat", k.max=20)
fviz_nbclust(tdat_wide[,-1], FUN = hcut, method = "silhouette", k.max=20)

#install.packages("cluster")
library(cluster)
gap_stat <- clusGap(tdat_wide[,-1], FUN = hcut, nstart = 25, K.max = 10, B = 100)
fviz_gap_stat(gap_stat)

# Compute distance matrix
res.dist <- dist(tdat_wide[,-1], method = "euclidean")
```

```{r  Calculate densities tumour type, message=FALSE, warning=FALSE, echo=FALSE}
tdat <-as.data.frame(table(all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=all.tumour_pat.roi$RoiID,
                                "Area"=all.tumour_pat.roi$Area_px_Core,
                                "Patient_ID"=all.tumour_pat.roi$Patient_ID,
                                "DX.name"=all.tumour_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Patient_ID) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)
  #t <-merge(t, clinical.data, by="Patient_ID")
  
  
#CAF type distribution over all patients
tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")
tdat_m <-as.matrix(tdat_wide)
rownames(tdat_m)<-tdat_wide$Patient_ID


hc <- hclust(dist(tdat_m[,-1]),"ward.D2")
#order Patients after clustering results
tdat_wide$Patient_ID <- factor(tdat_wide$Patient_ID, levels = labels(hc))
tdat$Patient_ID <- factor(tdat$Patient_ID, levels = labels(hc))
```

## Barplot showing absolute density per patietn coloured by tumour type together with hierarchical clustering tree coloured by patient metaclusters.
```{r Barolot tumour type densities, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, echo=FALSE}
#barolot (sqrt of density)
p <-ggplot(tdat,aes(y=sqrt(density),x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")

#barplot total density
p <-ggplot(tdat,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")#+theme(legend.position = "none")


#coloured dendrogram
ggdend <-hc%>% as.dendrogram() %>% dendextend::set("branches_k_color", k = 4) %>% as.ggdend()
p1 <-ggplot(ggdend, labels = F)+scale_x_continuous(expand = c(0, 0))+
  theme_void()
pg <-plot_grid(p1,p, align="v", ncol=1)
pg
#ggsave(plot=pg, file=file.path(plot_folder, paste("Barplot_Tumour-Category-Densities_HC_ordered_dendro.pdf")), width=12, height=6)
```
Add patient metacluster to data and plot clusters separately for better identification
```{r tumour type densities merge hc, message=FALSE, warning=FALSE, echo=FALSE}
clinical.data = clinical.data[!duplicated(clinical.data$Patient_ID),]

#cut dendrogram, add metacluster info to tdat
ct <-as.data.frame(cutree(hc, k=4))

colnames(ct)<-"metacluster"
ct$metacluster <-as.factor(ct$metacluster)
ct$Patient_ID <- rownames(ct)

tdat_wide_ct <-left_join(tdat_wide, ct, by="Patient_ID")
tdat_wide_ct <- left_join(tdat_wide_ct, clinical.data, by="Patient_ID")
table(tdat_wide_ct$metacluster)

#split barlot by metacluster
tdat_ct <-merge(tdat, ct, by="Patient_ID")
ggplot(tdat_ct,aes(y=density,x=Patient_ID, fill=Phenotype))+geom_bar(stat="identity")+scale_fill_tableau()+
    theme(legend.position="bottom")+
  facet_grid(~ metacluster, scales = "free", space = "free")
```

## Survival analysis over tumour type densities high low.  
- High > median
- Low < median
```{r  surv_dat tumour type densities high low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- surv_dat$normal+surv_dat$hypoxic

#high low
surv_dat$Density_G <- ifelse(surv_dat$total_density >summary(surv_dat$total_density)[3],"Density high","Density low")
surv_dat$normal_G <- ifelse(surv_dat$normal >summary(surv_dat$normal)[3],"normal high","normal low")
surv_dat$hypoxic_G <- ifelse(surv_dat$hypoxic >summary(surv_dat$hypoxic)[3],"hypoxic high","hypoxic low")
```

#CoxPH for tumour type density corrected for Stage, Grade and M
```{r  Coxph tumour type density corrected, echo=F, message=FALSE, warning=FALSE}
surv_dat
surv_dat<-surv_dat[!surv_dat$OS == 0, ]
time <- surv_dat$OS
status <- surv_dat$censoringOS
surv_dat$metacluster <-as.factor(surv_dat$metacluster)
surv_dat$Grade <-as.numeric(surv_dat$Grade)
rownames(surv_dat) <- surv_dat$Patient_ID
res.cox <- coxph(Surv(time, status) ~ metacluster, data = surv_dat)
res.cox


res.cox <- coxph(Surv(time, status) ~   normal + hypoxic + Stage+Grade+M.new, data =  surv_dat)

#res.cox <- coxph(Surv(time, status) ~ `normal+`+tpCAF+iCAF+mCAF+dCAF+vCAF+`SMA+ CAF`+`IDO+ CAF`+`dividing tpCAF`+`hypoxic CAF`+Grade, data =  surv_dat)
summary(res.cox)


td_res = tidy(res.cox, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
td_res <-td_res[td_res$term !="Grade"&
                  td_res$term !="Stage"&
                  td_res$term !="M.new",]

p <-ggplot(td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
#xlim(0,2)+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
#
plot(p)
```

## Not split by tumour type (AC and SQC together)
```{r  survival tumour type densities high low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
#surv <-c("metacluster")
surv <-c("hypoxic_G","normal_G","metacluster", "Density_G")
library(survminer)

for (i in surv){
  #for (k in unique(surv_dat$DX.name)){
    #dat <-surv_dat[surv_dat$DX.name==k,]
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
#km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## Split by tumour type (AC and SQC individually)
```{r  survival tumour type densities high low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Split
library(survminer)

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$Density_G <- ifelse(dat$total_density >summary(dat$total_density)[3],"Density high","Density low")
dat$normal_G <- ifelse(dat$normal >summary(dat$normal)[3],"normal high","normal low")
dat$hypoxic_G <- ifelse(dat$hypoxic >summary(dat$hypoxic)[3],"hypoxic high","hypoxic low")   
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()


#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```

## Lasso-regressed cox-ph model  
Including clinical parameters, patient metacluster and T cell categories (normal vs hypoxic) classified as high / low as well as the densities (continuous).   
```{r  coxph tumour type densities high-low, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=6, fig.width=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("Density_G","normal_G","hypoxic_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("normal_G","hypoxic_G","Density_G","metacluster",colnames(clinical.data),contains("censor"))) #)),"AG" #,NT,"Area","TMA_ImageID"

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients


active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tumour-Category_Density_high-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))


#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Survival tumour type densities high-mid-low     
- high    > 3rd quantile   
- mid     > 1st quantile < 3rd quantile   
- low     < 1st quantile   
```{r  surv_dat tumour type densities high mid low, message=FALSE, warning=FALSE, echo=FALSE}
surv_dat <-as.data.table(tdat_wide_ct)
surv_dat<- surv_dat[is.na(surv_dat$OS) !=T,]

# Censoring for OS (1 = event (death due to disease), 0 = censored)
surv_dat[,censoringOS := 0]
surv_dat[Ev.O == 1,censoringOS := 1]

# Censoring for disease free survival (1 = event (recurrence), 0 = censored)
surv_dat[ ,censoringDFS := 0]
surv_dat[ OS > DFS, censoringDFS := 1]
surv_dat$censoringDFS <-as.numeric(as.character(surv_dat$censoringDFS))

surv_dat <-surv_dat[surv_dat$DX.name=="Adenocarcinoma"|surv_dat$DX.name=="Squamous cell carcinoma",]
surv_dat <-surv_dat[!surv_dat$Patient_ID%in% neoadj_pat$Patient_ID,]

surv_dat$total_density <- surv_dat$normal+surv_dat$hypoxic

#High Med Low
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[5]] <-"Density high"
surv_dat$Density_G[surv_dat$total_density <summary(surv_dat$total_density)[2]] <-"Density low"
surv_dat$Density_G[surv_dat$total_density >summary(surv_dat$total_density)[2]& surv_dat$total_density<summary(surv_dat$total_density)[5]] <-"Density medium"

surv_dat$normal_G[surv_dat$normal >summary(surv_dat$normal)[5]] <-"normal high"
surv_dat$normal_G[surv_dat$normal <summary(surv_dat$normal)[2]] <-"normal low"
surv_dat$normal_G[surv_dat$normal >summary(surv_dat$normal)[2]& surv_dat$normal<summary(surv_dat$normal)[5]] <-"normal medium"

surv_dat$hypoxic_G[surv_dat$hypoxic >summary(surv_dat$hypoxic)[5]] <-"hypoxic high"
surv_dat$hypoxic_G[surv_dat$hypoxic <summary(surv_dat$hypoxic)[2]] <-"hypoxic low"
surv_dat$hypoxic_G[surv_dat$hypoxic >summary(surv_dat$hypoxic)[2]& surv_dat$hypoxic<summary(surv_dat$hypoxic)[5]] <-"hypoxic medium"

surv <-c("normal_G","hypoxic_G","Density_G","metacluster")
```

## tumour type densities split by tumour type
```{r  survival tumour type densities high-mid-low not split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
#Not split
surv <-c("normal_G","hypoxic_G","Density_G","metacluster")
library(survminer)

for (i in surv){
  dat <-surv_dat
  k="AC & SCC together"
#OS
km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)
#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

#print(p)
p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit

#disease free survival
km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit

pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
 #  abbr.colnames = FALSE, na = " "))
km_trt_fit
}
```

## tumour type densities split by tumour type
```{r  survival tumour type densities high-mid-low split, message=FALSE, warning=FALSE, echo=FALSE,out.width="50%", fig.height=4, fig.width=6}
surv <-c("normal_G","hypoxic_G","Density_G","metacluster")

for (i in surv){
  for (k in unique(surv_dat$DX.name)){
    dat <-surv_dat[surv_dat$DX.name==k,]
dat$Density_G[dat$total_density >summary(dat$total_density)[5]] <-"Density high"
dat$Density_G[dat$total_density <summary(dat$total_density)[2]] <-"Density low"
dat$Density_G[dat$total_density >summary(dat$total_density)[2]& dat$total_density<summary(dat$total_density)[5]] <-"Density medium"

dat$normal_G[dat$normal >summary(dat$normal)[5]] <-"normal high"
dat$normal_G[dat$normal <summary(dat$normal)[2]] <-"normal low"
dat$normal_G[dat$normal >summary(dat$normal)[2]& dat$normal<summary(dat$normal)[5]] <-"normal medium"

dat$hypoxic_G[dat$hypoxic >summary(dat$hypoxic)[5]] <-"hypoxic high"
dat$hypoxic_G[dat$hypoxic <summary(dat$hypoxic)[2]] <-"hypoxic low"
dat$hypoxic_G[dat$hypoxic >summary(dat$hypoxic)[2]& dat$hypoxic<summary(dat$hypoxic)[5]] <-"hypoxic medium"
  
#OS
pw<-pairwise_survdiff(Surv(OS, censoringOS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(OS, censoringOS) ~ dat[[i]], data=dat)  
km_trt_fit
p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = F,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("OS over",i,"per",k))
 # facet_grid(.~surv_dat[[i]])
print(p)


#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-med-low_OS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#km_trt_fit
#print(p)

#disease free survival
pw<-pairwise_survdiff(Surv(DFS, censoringDFS) ~ metacluster,data = dat)
#print(pw)
#print(symnum(pw$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
#   symbols = c("****", "***", "**", "*", "+", "."),
#   abbr.colnames = FALSE, na = " "))

km_trt_fit <- survfit(Surv(DFS, censoringDFS) ~ dat[[i]], data=dat)      
km_trt_fit

p <-ggsurvplot(km_trt_fit,
               data=dat,
               conf.int = T,
               pval = T,
               #conf.int.style = "step",
               surv.median.line = "hv",
               xlim=c(0,5000))+ # Specify median survival,
               #legend.title = paste(i),
               #legend.labs=c("Grade 1","Grade 2","Grade 3"),
               #legend.labs=c(paste(i,sort(unique(surv_dat[[i]][!is.na(surv_dat[[i]])])))))+
  labs(title=paste("DFS over",i,"per",k))#+
 # facet_grid(.~surv_dat[[i]])
print(p)

#pdf(file=file.path(plot_folder, paste0("Survival_Density-tumour_type_hi-med-low_DFS over",i,"_per_",k,".pdf")))
#print(p,newpage=FALSE)
#dev.off()

p <-p$plot+
  theme_bw()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  #facet_wrap(~strata)+
  theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank())
#print(p)
#km_trt_fit
  }
}
```



## Lasso-regressed cox-ph model   
including clinical parameters, patient metacluster and T cell categories (normal vs hypoxic) classified as high mid low as well as the densities (continuous).   
```{r coxph tumour type densities high-mid-low, message=FALSE, warning=FALSE, echo=FALSE, out.width="50%", fig.width=6, fig.height=6}
surv_dat$metacluster <-as.numeric(surv_dat$metacluster)
fd_test <- fastDummies::dummy_cols(surv_dat, select_columns = c("Density_G","normal_G","hypoxic_G","Typ","Grade","T.new","N","M.new","Stage","Smok","Gender", "metacluster")) #

fd_test<-fd_test[!fd_test$OS == 0, ]
time <- fd_test$OS
status <- fd_test$censoringOS

rownames(fd_test) <- fd_test$Patient_ID

fd_test <-fd_test %>%select(-c("normal_G","hypoxic_G","Density_G","metacluster",colnames(clinical.data),contains("censor")))

fd_test[is.na(fd_test)] <- 0
fd_test_matrix <-data.matrix(fd_test)


#determine lambda
cv.fit <- cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
fit <- glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
#plot(cv.fit)
#cv.fit$lambda.min

#loop to determine optimal lambda for lasso regression
MSEs <- NULL
for (i in 1:100){
                 cv <-  cv.glmnet(fd_test_matrix, Surv(time, status), family = "cox", maxit = 1000)
                 MSEs <- cbind(MSEs, cv$cvm)
             }
  rownames(MSEs) <- cv$lambda
  lambda.min <- as.numeric(names(which.min(rowMeans(MSEs))))

#coef(fit, s = lambda.min)
Coefficients <- coef(fit, s = lambda.min)
Active.Index <- which(Coefficients != 0)
Active.Coefficients <- Coefficients[Active.Index]
#Coefficients

active_coefficients <- Coefficients[,1] != 0
x <- fd_test_matrix[,active_coefficients]
x <- data.frame(x)

vct <-colnames(x)
coxdf<-coxph(Surv(time,status)~.,data=x)

td_res = tidy(coxdf, exponentiate = TRUE)
td_res = td_res %>% arrange(estimate) %>% mutate(term = factor(term,levels = term))
td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.high = estimate + std.error)

td_res<-td_res %>%
  rowwise() %>%
  dplyr::mutate(conf.low = estimate - std.error)
td_res<-td_res[!td_res$std.error == 0, ]
na_td_res <- td_res[-grep("NA",td_res$term),]


p <-ggplot(na_td_res, aes(x = term, y = estimate,color =p.value < 0.05)) +
geom_point(size = 4) +
geom_errorbar(aes(ymax = conf.high, ymin = conf.low))+
scale_colour_viridis_d("p value < 0.05")+
coord_flip()+
geom_hline(yintercept = 1,color = 'black')+
theme_bw()+
scale_colour_tableau()
p
#ggsave(file=file.path(plot_folder, paste("CoxPH_Tumour-Category_Density_high-mid-low_Lasso.pdf")), width=10, height=10, plot=p)

ggforest(coxdf, data =  (fd_test %>% select(-contains("NA"))))

#pdf(file=file.path(plot_folder,paste("COX_selectedFactors_glmnet.pdf")))
#ggforest(coxdf, data = fd_test)
#dev.off()
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables without neoadjuvant treated patients.
```{r  KW W tumour type density WO neo,warning=FALSE, message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

all.tumour_pat.roi$NeoAdj <- ifelse(all.tumour_pat.roi$Chemo==1 |all.tumour_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=all.tumour_pat.roi$RoiID,
                                "Area"=all.tumour_pat.roi$Area_px_Core,
                                "Patient_ID"=all.tumour_pat.roi$Patient_ID,
                                "DX.name"=all.tumour_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$normal + tdat_wide$hypoxic
tdat <-tdat_wide %>% pivot_longer(cols=c("normal","hypoxic","total_density"), names_to = "Phenotype",values_to = "density")

#remove
tdat <-tdat[!tdat$Patient_ID%in% neoadj_pat$Patient_ID,]

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)+
      theme(legend.position = "none")
    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=3) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
 # pdf(file=file.path(plot_folder, paste0("KW_W_Tumour-Category_Densities_WOneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

## Kruskal Wallis / Wilcoxon comparison between groups of clincial variables with neoadjuvant treated patients.
```{r  KW Wilcox tumour type density W neo,warning=F,message=FALSE, echo=FALSE, fig.width=12, fig.height=6}
#define variables to be tested here
categories <- c( "Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender")

all.tumour_pat.roi$NeoAdj <- ifelse(all.tumour_pat.roi$Chemo==1 |all.tumour_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=all.tumour_pat.roi$RoiID,
                                "Area"=all.tumour_pat.roi$Area_px_Core,
                                "Patient_ID"=all.tumour_pat.roi$Patient_ID,
                                "DX.name"=all.tumour_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$normal + tdat_wide$hypoxic
tdat <-tdat_wide %>% pivot_longer(cols=c("normal","hypoxic","total_density"), names_to = "Phenotype",values_to = "density")

tdat_c <-merge(tdat, clinical.data, by="Patient_ID")
categories <- c("Dist.Met", "T.new","M.new","N","Relapse","DX.name","Stage","Grade","Gender","Chemo","Radio","NeoAdj")
plot_list <- list()
for (i in (categories)) {
  for (j in unique(tdat_c$Phenotype)){
    tdat<-tdat_c
    tdat[[i]] <-as.factor(tdat[[i]])
    tdat <-tdat %>% drop_na(i)
    tdat$i <-as.factor(tdat[[i]])
    tdat$Phenotype <-as.factor(tdat$Phenotype)
    tdat$Phenotype <-droplevels(tdat$Phenotype)

    tdat <-subset(tdat, Phenotype==j)

    if (length(unique(tdat[[i]])) > 2)
      {
    #calculate p values, add position for plot
    stat.test <- tdat%>%
     group_by(Phenotype) %>%
     dunn_test(density~i,p.adjust.method = "bonferroni")
    stat.test<-stat.test %>% select(-.y., -statistic)
    stat.test <- stat.test %>% add_xy_position(x = i)
   
    #create plot list
    plot_list[[j]]<-
      ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
              geom_violin(width=0.5) +
    geom_boxplot(width=0.3, color="grey", alpha=0.2)+
      #geom_boxplot()+
      geom_point()+
      facet_wrap(~Phenotype, scales="free", ncol=3, strip.position="top")+
      scale_color_viridis_d()+
      theme_bw()+
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
      labs( y="Mean patient density")+
      theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")+
      stat_pvalue_manual(stat.test, label = "p.adj.signif", tip.length = 0, hide.ns = T)+
      theme(legend.position = "none")
    }
    
     else{
      
        tdat<-tdat_c
        tdat[[i]] <-as.factor(tdat[[i]])
        tdat <-tdat %>% drop_na(i)
        tdat$Phenotype <-factor(tdat$Phenotype)
        tdat$Phenotype <-droplevels(tdat$Phenotype)
        tdat <-tdat %>% drop_na(i)

        tdat <-subset(tdat, Phenotype==j)
          detach("package:dplyr", unload=TRUE)
          library(dplyr)
      pvalues <- tdat %>% 
      group_by(Phenotype) %>% 
      summarise(p=wilcox.test(tdat$density~tdat[[i]], paired=F)$p.value)
      tdat <- merge(tdat, pvalues, by.x = "Phenotype", by.y ="Phenotype", all.x = TRUE)
      tdat$p.wt <- paste0('p=',round(tdat$p, digits=3))
    
      #Plot list, all js together over i
      plot_list[[j]]<-
        ggplot(tdat, aes(x= .data[[i]], y = density, colour=.data[[i]]))+
        geom_violin(width=0.5) +
        geom_boxplot(width=0.3, color="grey", alpha=0.2)+
        #geom_boxplot()+
        geom_point()+
        facet_wrap(~Phenotype+p.wt, scales="free", ncol=3, strip.position="top")+
        #ggtitle("paired")+
        theme(axis.title.x=element_text("Phenotype"))+
      
        scale_color_viridis_d()+
        theme_bw()+
        theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
        labs(y="Mean patient density", fill="Area_Phenotype")+
        theme(axis.text.x = element_text(angle=45, hjust=1), axis.ticks.x = element_blank(),legend.position = "none")
    }  
 
    }

  #plot
  gridExtra::grid.arrange(grobs = plot_list, ncol=3) #, ncol=round(length(unique(t$Phenotype)))
  
  #save out in individual pdfs for each variable
#  pdf(file=file.path(plot_folder, paste0("KW_W_Tumour-Category_Densities_Wneo",i,".pdf")), width=12, height=6)
 # gridExtra::grid.arrange(grobs = plot_list, ncol=3)
#  dev.off()
}
```

```{r  Correlation tumour type densities with clinical parameters, eval=F, echo=T, message=FALSE, warning=FALSE, echo=FALSE}
all.tumour_pat.roi$NeoAdj <- ifelse(all.tumour_pat.roi$Chemo==1 |all.tumour_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
all.tumour_pat.roi$NeoAdj <- ifelse(all.tumour_pat.roi$Chemo==1 |all.tumour_pat.roi$Radio==1,"NeoAdjuvantTherapy","NoNeoAdjuvantTherapy")
tdat <-as.data.frame(table(all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$RoiID,
                           all.tumour_pat.roi[,all.tumour_pat.roi$Patient_ID!="Control"&
                                          all.tumour_pat.roi$DX.name=="Adenocarcinoma"|
                                          all.tumour_pat.roi$DX.name=="Squamous cell carcinoma"]$tumour_type))
colnames(tdat)<-c("RoiID","Phenotype","n")

da <-dplyr::distinct(data.frame("RoiID"=all.tumour_pat.roi$RoiID,
                                "Area"=all.tumour_pat.roi$Area_px_Core,
                                "Patient_ID"=all.tumour_pat.roi$Patient_ID,
                                "DX.name"=all.tumour_pat.roi$DX.name))

  tdat <-merge(tdat,da,by="RoiID")
  tdat <-tdat[tdat$DX.name=="Adenocarcinoma" | tdat$DX.name=="Squamous cell carcinoma",]
  tdat <-tdat %>%
    #dplyr::count(Type, Area,Sample) %>%
    dplyr::group_by(RoiID) %>%
    dplyr::mutate(density = n / Area*1000000)
  range(tdat$density)

  tdat <-tdat %>% 
  group_by(Patient_ID,Phenotype) %>%
  summarise(across(-c(RoiID,n,Area,DX.name), sum, na.rm = TRUE))
  range(tdat$density)

tdat_wide <- pivot_wider(tdat,id_cols=c("Patient_ID"),names_from = "Phenotype", values_from ="density",names_prefix = "")

tdat_wide$total_density <- tdat_wide$normal + tdat_wide$hypoxic

tdat_wide <-merge(tdat_wide, clinical.data, by="Patient_ID")
tdat_wide_as <-tdat_wide[tdat_wide$DX.name=="Adenocarcinoma" | tdat_wide$DX.name=="Squamous cell carcinoma",]
tdat_wide_as$LN.Met <-ifelse(tdat_wide_as$LN.Met=="LN Metastases", 1,0)
tdat_wide_as$Dist.Met <-ifelse(tdat_wide_as$Dist.Met=="Dist. Metastases", 1,0)

for(i in c.param){
  print(i)
  print(cor.test(tdat_wide_as$normal,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$hypoxic,tdat_wide_as[[i]], method="spearman", exact=F))
  print(cor.test(tdat_wide_as$total_density,tdat_wide_as[[i]], method="spearman", exact=F))
}
```
